# Cluster specific filters for IBD and selection


# Create files with list of samples to exclude for each cluster

Done with ordination_plots.R

# Filtering MAF by clusters

 - genotype and sample missingness filters
 - MAF 0.05 filters for the clusters independently
    - 'Exclude' samples are generated in the fastLMM Rmd 
 - merge outputs from MAF filters 
 - run genotype missingness filter again - as some variants might not overlap because of the application to the MAF filters to the clusters

```{R,eval=F}
mkdir cluster_specific_maf_filters/high_quality_set

plink --file Pk --geno 0.20 --make-bed --out cluster_specific_maf_filters/high_quality_set/cleaned
plink --bfile cluster_specific_maf_filters/high_quality_set/cleaned --recode --out cluster_specific_maf_filters/high_quality_set/cleaned
plink --file cluster_specific_maf_filters/high_quality_set/cleaned --mind 0.05 --make-bed --out cluster_specific_maf_filters/high_quality_set/cleaned
plink --bfile cluster_specific_maf_filters/high_quality_set/cleaned --recode --out cluster_specific_maf_filters/high_quality_set/cleaned
plink --file cluster_specific_maf_filters/high_quality_set/cleaned --geno 0.05 --make-bed --out cluster_specific_maf_filters/high_quality_set/cleaned

plink --bfile cleaned --recode  --remove Exclude_for_Mf.tsv --out cluster_specific_maf_filters/high_quality_set/Mf
plink --file cluster_specific_maf_filters/high_quality_set/Mf --maf 0.05 --make-bed --out cluster_specific_maf_filters/high_quality_set/Mf

plink --bfile cleaned --recode  --remove Exclude_for_Mn.tsv --out cluster_specific_maf_filters/high_quality_set/Mn
plink --file cluster_specific_maf_filters/high_quality_set/Mn --maf 0.05 --make-bed --out cluster_specific_maf_filters/high_quality_set/Mn

plink --bfile cleaned --recode  --remove Exclude_for_Pen.tsv --out cluster_specific_maf_filters/high_quality_set/Pen
plink --file cluster_specific_maf_filters/high_quality_set/Pen --maf 0.05 --make-bed --out cluster_specific_maf_filters/high_quality_set/Pen
```

# Wrangling data for IBD analysis 

```{bash,eval=F}
awk '{print $1}' cluster_specific_maf_filters/high_quality_set/Mf.fam > exclude_tmp1.txt
bcftools view Consensus_SNPs_subset_no_MOI.vcf.gz | grep '#CHROM' | tr '\t' '\n' | sed -n '/ERR2214837/,$p' > exclude_tmp2.txt
fgrep -v -f exclude_tmp1.txt exclude_tmp2.txt > cluster_specific_maf_filters/high_quality_set/exclude_tmp3.txt
cat cluster_specific_maf_filters/high_quality_set/exclude_tmp3.txt dups_from_naming_error.txt lab_isolates.txt > cluster_specific_maf_filters/high_quality_set/exclude_for_Mf.txt

awk '{print $2}' cluster_specific_maf_filters/high_quality_set/Mf.bim | sed 's/:/\t/g' | awk '{print $1"\t"$2}' | uniq -u > cluster_specific_maf_filters/high_quality_set/Mf_grep_patterns.tsv

awk '{print $1}' cluster_specific_maf_filters/high_quality_set/Mn.fam > exclude_tmp1.txt
bcftools view Consensus_SNPs_subset_no_MOI.vcf.gz | grep '#CHROM' | tr '\t' '\n' | sed -n '/ERR2214837/,$p' > exclude_tmp2.txt
fgrep -v -f exclude_tmp1.txt exclude_tmp2.txt > cluster_specific_maf_filters/high_quality_set/exclude_tmp3.txt
cat cluster_specific_maf_filters/high_quality_set/exclude_tmp3.txt dups_from_naming_error.txt lab_isolates.txt > cluster_specific_maf_filters/high_quality_set/exclude_for_Mn.txt

awk '{print $2}' cluster_specific_maf_filters/high_quality_set/Mn.bim | sed 's/:/\t/g' | awk '{print $1"\t"$2}' | uniq -u > cluster_specific_maf_filters/high_quality_set/Mn_grep_patterns.tsv

awk '{print $1}' cluster_specific_maf_filters/high_quality_set/Pen.fam > exclude_tmp1.txt
bcftools view Consensus_SNPs_subset_no_MOI.vcf.gz | grep '#CHROM' | tr '\t' '\n' | sed -n '/ERR2214837/,$p' > exclude_tmp2.txt
fgrep -v -f exclude_tmp1.txt exclude_tmp2.txt > cluster_specific_maf_filters/high_quality_set/exclude_tmp3.txt
cat cluster_specific_maf_filters/high_quality_set/exclude_tmp3.txt dups_from_naming_error.txt lab_isolates.txt > cluster_specific_maf_filters/high_quality_set/exclude_for_Pen.txt

awk '{print $2}' cluster_specific_maf_filters/high_quality_set/Pen.bim | sed 's/:/\t/g' | awk '{print $1"\t"$2}' | uniq -u > cluster_specific_maf_filters/high_quality_set/Pen_grep_patterns.tsv

rm -f exclude_tmp1.txt exclude_tmp2.txt cluster_specific_maf_filters/high_quality_set/*tmp*

```

# Create genotype file for each

Rscript /g/data/pq84/malaria/Parasite_and_human_genetic_risk_factors_for_Pk_malaria/scripts/05_Analyses/Population_genetics/hmmIBD/genotype_clusters.R

```{r,eval=F}
# Packages
library(dplyr)
library(readr)
library(stringr)
library(data.table)

# Read in and wrangle VCF

create_IBD_genotype <- function(FILE_PATH){
    read_tsv(FILE_PATH, col_names = c("CHROM", "POS")) %>%
        mutate(POS = as.numeric(POS)) %>%
        left_join(
                read_table("hmmIBD/Consensus_SNPs_subset_no_MOI.vcf", skip = 72) %>%
                rename("CHROM" = `#CHROM`) %>% # this first section selects only variants that passed the PLINK filters we applied
                mutate(POS = as.numeric(POS)) 
        ) %>%
        mutate(CHROM = str_remove(CHROM, "ordered_PKNH_"),
        CHROM = str_remove(CHROM, "_v2")) %>%
        mutate_at(c(10:ncol(.)), ~str_remove(., ":.*")) %>% 
        mutate_at(c(10:ncol(.)), ~ifelse(. %like% "1/1" | . %like% "1/0" | . %like% "0/1" , "1", .)) %>% 
        mutate_at(c(10:ncol(.)), ~ifelse(. %like% "2/2" | . %like% "2/0" | . %like% "0/2" , "2", .)) %>% 
        mutate_at(c(10:ncol(.)), ~ifelse(. %like% "3/3" | . %like% "3/0" | . %like% "0/3" , "3", .)) %>% 
        mutate_at(c(10:ncol(.)), ~ifelse(. %like% "4/4" | . %like% "4/0" | . %like% "0/4" , "4", .)) %>% 
        mutate_at(c(10:ncol(.)), ~ifelse(. %like% "0/0" , "0", .)) %>% # 0 for reference allele
        mutate_at(c(10:ncol(.)), ~ifelse(. %like% "./." , "-1", .)) %>% # for missing 
        select(-c(3:9)) %>%
        arrange(CHROM, POS) %>%
        write_tsv(paste0(str_replace(FILE_PATH, "_grep_patterns.tsv", "_hmmIBD.tsv")))
}

list_of_files <- list.files(path = "cluster_specific_maf_filters/high_quality_set", pattern = "patterns.tsv", full.names = TRUE) 

for (i in list_of_files){
    create_IBD_genotype(paste0(i))
}
```

# Run hmmIBD

```{R,eval=F}
hmmIBD -i hmmIBD/cluster_spec/Mf_hmmIBD.tsv -b cluster_specific_maf_filters/high_quality_set/exclude_for_Mf.txt -o hmmIBD/cluster_spec/Mf
hmmIBD -i hmmIBD/cluster_spec/Mn_hmmIBD.tsv -b cluster_specific_maf_filters/high_quality_set/exclude_for_Mn.txt -o hmmIBD/cluster_spec/Mn
hmmIBD -i hmmIBD/cluster_spec/Pen_hmmIBD.tsv -b cluster_specific_maf_filters/high_quality_set/exclude_for_Pen.txt -o hmmIBD/cluster_spec/Pen
```


# Selection



# Variant selection - [rehh](https://cran.r-project.org/web/packages/rehh/vignettes/rehh.html#LoadData)

Use clonal_dups.txt to filter out clonal samples - messes with alogrithm.


Since the whole file is read in for rehh, even though you can only look at a single chr at a time, it is advisable to split large data sets into chromosome-specific files.
In addition, computing Rsb requires two data frames containing each the inES statistics of one population as obtained by the scan_hh() function - THUS - need to create separate VCF files for the different clusters that we want to compare.

Steps:
    - filter VCF using high quality GWAS type set of variants produced through plink (high_quality_set)
    - remove samples not in high quality set, dups_from_naming_error.txt, lab_isolates.txt, clonal_dups.txt
    - covert missing data to reference - otherwise they're filtered out by rehh
    - create a lsit of the chromosomes
    - get a list of samples to be included (-S) in the rehh analysis (Mn of Mf)
    - subset VCF into a VCF for individual chromosomes
    - subset to clusters

**Need to ammend the sn and xl line manually based on outputs of "cat"**


Pulling exclude samples from above and adding to clonal duplicates


```{bash}
mkdir selection/cluster_spec/Mf

awk '{print $2}' cluster_specific_maf_filters/high_quality_set/Mf.bim | sed 's/:/\t/g' | awk '{print $1 "\t" $2}' > selection/cluster_spec/Mf/snp_filter.txt
cat cluster_specific_maf_filters/high_quality_set/exclude_for_Mf.txt clonal_dups.txt > selection/cluster_spec/Mf/exclude.txt
bcftools view --force-samples -R selection/cluster_spec/Mf/snp_filter.txt -S ^selection/cluster_spec/Mf/exclude.txt Consensus_SNPs_subset_no_MOI.vcf.gz | bcftools +missing2ref -o selection/cluster_spec/Mf/selection.vcf.gz
bcftools index selection/cluster_spec/Mf/selection.vcf.gz
bcftools view selection/cluster_spec/Mf/selection.vcf.gz | grep '##contig=<ID=ordered' | sed 's/##contig=<ID=//g' | sed 's/,.*//g' > selection/cluster_spec/chromosome.txt

for i in $(cat selection/cluster_spec/chromosome.txt)
    do 
        bcftools view -r $i -o selection/cluster_spec/Mf/${i}.vcf.gz selection/cluster_spec/Mf/selection.vcf.gz
    done 

mkdir selection/cluster_spec/Mn

awk '{print $2}' cluster_specific_maf_filters/high_quality_set/Mn.bim | sed 's/:/\t/g' | awk '{print $1 "\t" $2}' > selection/cluster_spec/Mn/snp_filter.txt
cat cluster_specific_maf_filters/high_quality_set/exclude_for_Mn.txt clonal_dups.txt > selection/cluster_spec/Mn/exclude.txt
bcftools view --force-samples -R selection/cluster_spec/Mn/snp_filter.txt -S ^selection/cluster_spec/Mn/exclude.txt Consensus_SNPs_subset_no_MOI.vcf.gz | bcftools +missing2ref -o selection/cluster_spec/Mn/selection.vcf.gz
bcftools index selection/cluster_spec/Mn/selection.vcf.gz
bcftools view selection/cluster_spec/Mn/selection.vcf.gz | grep '##contig=<ID=ordered' | sed 's/##contig=<ID=//g' | sed 's/,.*//g' > selection/cluster_spec/chromosome.txt

for i in $(cat selection/cluster_spec/chromosome.txt)
    do 
        bcftools view -r $i -o selection/cluster_spec/Mn/${i}.vcf.gz selection/cluster_spec/Mn/selection.vcf.gz
    done 

mkdir selection/cluster_spec/Pen

awk '{print $2}' cluster_specific_maf_filters/high_quality_set/Pen.bim | sed 's/:/\t/g' | awk '{print $1 "\t" $2}' > selection/cluster_spec/Pen/snp_filter.txt
cat cluster_specific_maf_filters/high_quality_set/exclude_for_Pen.txt clonal_dups.txt > selection/cluster_spec/Pen/exclude.txt
bcftools view --force-samples -R selection/cluster_spec/Pen/snp_filter.txt -S ^selection/cluster_spec/Pen/exclude.txt Consensus_SNPs_subset_no_MOI.vcf.gz | bcftools +missing2ref -o selection/cluster_spec/Pen/selection.vcf.gz
bcftools index selection/cluster_spec/Pen/selection.vcf.gz
bcftools view selection/cluster_spec/Pen/selection.vcf.gz | grep '##contig=<ID=ordered' | sed 's/##contig=<ID=//g' | sed 's/,.*//g' > selection/cluster_spec/chromosome.txt

for i in $(cat selection/cluster_spec/chromosome.txt)
    do 
        bcftools view -r $i -o selection/cluster_spec/Pen/${i}.vcf.gz selection/cluster_spec/Pen/selection.vcf.gz
    done 
```


# Selection with rehh 

Rscript /g/data/pq84/malaria/Parasite_and_human_genetic_risk_factors_for_Pk_malaria/scripts/05_Analyses/Population_genetics/rehh.R


```{r,eval=F}
# Packages
library(rehh)
library(tidyverse)
library(data.table)
library(R.utils)

# WHOLE GENOME iHS

## need to read in chr files independently and build up an object resulting from the scan_hh outputs
### first create an empty df to be appended in the loop and a list of chr names
### the loop then creates the vcf file name (hap_file), and reads in the file to hh, which is then parsed into the scan_hh function to produce an object i that is appended to the growing wgscan object
### the resulting wgscan produces the same results as the scan_hh object above
### this can then be parsed into ihh2ihs

chromsomes_files <- c("ordered_PKNH_01_v2", "ordered_PKNH_02_v2", "ordered_PKNH_03_v2", "ordered_PKNH_04_v2", "ordered_PKNH_05_v2",
    "ordered_PKNH_06_v2", "ordered_PKNH_07_v2", "ordered_PKNH_08_v2", "ordered_PKNH_09_v2", "ordered_PKNH_10_v2", "ordered_PKNH_11_v2",
    "ordered_PKNH_12_v2", "ordered_PKNH_13_v2", "ordered_PKNH_14_v2")

wgscan <- tibble("CHR" = NA, "POSITION" = NA, "FREQ_A" = NA, "FREQ_D" = NA, "NHAPLO_A" = NA, "NHAPLO_D" = NA, "IHH_A" = NA, "IHH_D" = NA, "IES" = NA, "INES" = NA)

for(i in chromsomes_files) {
    # haplotype file name for each chromosome
    hap_file = paste(i, ".vcf.gz", sep = "") # filename pattern
    # create internal representation
    hh <- data2haplohh(hap_file = hap_file, 
                    chr.name = i, 
                    polarize_vcf = FALSE, # if the AA key is absent
                    min_perc_geno.mrk = 100, # discard markers genotyped on < 100% of haplotypes
                    min_maf = 0.05, # discard markers with a minor allele frequency of < 0.05)
                    vcf_reader = "data.table") # use this package
    scan <- scan_hh(hh) # perform scan on a single chromosome (calculate iHH/iES values)
    wgscan <- wgscan %>% # append the wgscan object every round with additional data
    rbind(scan)
}

wgs_ihs <- wgscan %>%
    na.omit() %>%
    ihh2ihs(., min_maf = 0.05, # default
            freqbin = 0.025 # default
            )

### Output is a list with two elements:
### ihs: a data frame with iHS and corresponding p-value piHS (p-value in a −log10 scale assuming the iHS are normally distributed under the neutral hypothesis)
### frequency.class: a data frame summarizing the derived allele frequency bins used for standardization and mean and standard deviation of the un-standardized values
wgs_ihs$ihs
wgs_ihs$frequency.class


# Scan for candidate regions 
## First update iHS table so that its compatible
wgs_ihs_2 <- wgs_ihs
wgs_ihs_2$ihs <- wgs_ihs_2$ihs %>%
    mutate(CHR = str_remove(CHR, "ordered_PKNH_")) %>%
    mutate(CHR = str_remove(CHR, "_v2")) %>%
    mutate(CHR = as.numeric(CHR)) %>%
    as.data.frame()

candidate_regions <- calc_candidate_regions(wgs_ihs_2,
                                 threshold = 4,
                                 pval = TRUE,
                                 window_size = 10000,
                                 overlap = 1000,
                                 min_n_extr_mrk = 3) %>%
                                 add_column(Stat = "iHS") 

## Save candidate regions for all stats
candidate_regions %>%
  write_tsv("candidate_regions_iHS.tsv")

selection_plots <- function(DATA, MODEL){
  plot_data <- DATA %>%
    na.omit() %>%
    mutate(CHR = str_remove(CHR, "ordered_PKNH_")) %>%
    mutate(CHR = str_remove(CHR, "_v2")) %>%
    mutate(CHR = as.numeric(CHR)) %>%
    arrange(CHR, POSITION) %>%
    mutate(POS = 1:nrow(.)) %>%
    mutate(CHR = as.factor(CHR))

  x_axis <- plot_data %>%
    group_by(CHR) %>%
    summarise(POS = median(POS)) %>%
    mutate(CHR = str_remove(CHR, "^0"))

  hline <- plot_data %>% 
    summarise(enframe(quantile(!!sym(MODEL), c(0.001, 0.5, 0.999)), "quantile", MODEL))

  selection_plot <- plot_data %>%
    ggplot(aes(x = POS, y = !!sym(MODEL), colour = CHR)) +
    geom_point() + 
    theme(legend.position = "none") +
    scale_colour_manual(values = c("#39568CFF", "#29AF7FFF", "#39568CFF", "#29AF7FFF", "#39568CFF", "#29AF7FFF", "#39568CFF", "#29AF7FFF", "#39568CFF", "#29AF7FFF", "#39568CFF", "#29AF7FFF", "#39568CFF", "#29AF7FFF")) +
    scale_x_continuous(breaks = x_axis$POS, labels = x_axis$CHR) +
    xlab("Position") +
    ylab(MODEL) +
    geom_hline(yintercept = c(as.numeric(hline[1,2]), as.numeric(hline[3,2])), linetype="dotted")

  ggsave(paste0(MODEL, ".png"), dpi = 300, width = 16, selection_plot)

  selection_plot <- plot_data %>%
      ggplot(aes(x = POS, y = LOGPVALUE, colour = CHR)) +
      geom_jitter() + 
      theme(legend.position = "none") +
      scale_x_continuous(breaks = x_axis$POS, labels = x_axis$CHR) +
      scale_colour_manual(values = c("#39568CFF", "#29AF7FFF", "#39568CFF", "#29AF7FFF", "#39568CFF", "#29AF7FFF", "#39568CFF", "#29AF7FFF", "#39568CFF", "#29AF7FFF", "#39568CFF", "#29AF7FFF", "#39568CFF", "#29AF7FFF")) +
      xlab("Chromosomes") +
      ylab(expression(MODEL ~ ~-log[10](italic(p))))

  ggsave(paste0(MODEL, "_pvalue.png"), dpi = 300, width = 16, selection_plot)
}

selection_plots(wgs_ihs$ihs, "IHS")
```