# Cluster specific filters for IBD and selection

## Required tools/libraries for Analyses

```{bash,eval=F}
start_pbs_big_boy
module load bcftools/1.12
module load java/jdk-8.40 
module load R/4.3.1 
export R_LIBS=/home/588/jw1542/.local/lib/R_4.3:$R_LIBS
export PATH=$PATH:/g/data/pq84/bin/plink2/
source /g/data/pq84/python-3.9.2/bin/activate
export PATH=$PATH:/g/data/pq84/bin/hmmIBD/  

FASTA="/g/data/pq84/malaria/Parasite_and_human_genetic_risk_factors_for_Pk_malaria/data/ref_genomes/PKA1H1/fasta"
VCF="/g/data/pq84/malaria/pk_global_popgen/snakemake_pipeline/variant_calling/output/calling/consensus/filtered"
GATK="/g/data/pq84/bin/GenomeAnalysisTK-3.8-1-0/GenomeAnalysisTK.jar"
REGIONS_TO_MASK="/g/data/pq84/malaria/Parasite_and_human_genetic_risk_factors_for_Pk_malaria/data/ref_genomes/regions_to_mask.list"

cd /g/data/pq84/malaria/pk_global_popgen/outputs/merged
```

# Create files with list of samples to exclude for each cluster

Done with ordination_plots.R

# Filtering MAF by clusters

 - genotype and sample missingness filters
 - MAF 0.05 filters for the clusters independently
    - 'Exclude' samples are generated in the fastLMM Rmd 
 - merge outputs from MAF filters 
 - run genotype missingness filter again - as some variants might not overlap because of the application to the MAF filters to the clusters

```{bash,eval=F}
mkdir cluster_specific_maf_filters

plink --bfile Pk --recode --out Pk
plink --file Pk --geno 0.20 --make-bed --out cluster_specific_maf_filters/cleaned
plink --bfile cluster_specific_maf_filters/cleaned --recode --out cluster_specific_maf_filters/cleaned
plink --file cluster_specific_maf_filters/cleaned --mind 0.05 --make-bed --out cluster_specific_maf_filters/cleaned
plink --bfile cluster_specific_maf_filters/cleaned --recode --out cluster_specific_maf_filters/cleaned
plink --file cluster_specific_maf_filters/cleaned --geno 0.1 --make-bed --out cluster_specific_maf_filters/cleaned

plink --bfile cluster_specific_maf_filters/cleaned --recode  --remove outputs/Exclude_for_Mf.tsv --out cluster_specific_maf_filters/Mf
plink --file cluster_specific_maf_filters/Mf --maf 0.01 --make-bed --out cluster_specific_maf_filters/Mf

plink --bfile cluster_specific_maf_filters/cleaned --recode  --remove outputs/Exclude_for_Mn.tsv --out cluster_specific_maf_filters/Mn
plink --file cluster_specific_maf_filters/Mn --mac 2 --make-bed --out cluster_specific_maf_filters/Mn

plink --bfile cluster_specific_maf_filters/cleaned --recode  --remove outputs/Exclude_for_Pen.tsv --out cluster_specific_maf_filters/Pen
plink --file cluster_specific_maf_filters/Pen --mac 2 --make-bed --out cluster_specific_maf_filters/Pen
```

# Wrangling data for IBD analysis 


```{R,eval=F}
awk '{print $1}' cluster_specific_maf_filters/Mf.fam > exclude_tmp1.txt
bcftools view -h Consensus_SNPs_subset_no_MOI.vcf.gz | grep '#CHROM' | tr '\t' '\n' | sed -n '/FORMAT/,$p' | tail -n +2 > exclude_tmp2.txt
fgrep -v -f exclude_tmp1.txt exclude_tmp2.txt > cluster_specific_maf_filters/exclude_tmp3.txt
cat cluster_specific_maf_filters/exclude_tmp3.txt <(awk '{print $1}' Pk.dups) lab_isolates.txt | sort | uniq > cluster_specific_maf_filters/hmmIBD/exclude_for_Mf.txt

awk '{print $2}' cluster_specific_maf_filters/Mf.bim | sed 's/:/\t/g' | awk '{print $1"\t"$2}' | uniq -u > cluster_specific_maf_filters/hmmIBD/Mf_grep_patterns.tsv

awk '{print $1}' cluster_specific_maf_filters/Mn.fam > exclude_tmp1.txt
bcftools view -h Consensus_SNPs_subset_no_MOI.vcf.gz | grep '#CHROM' | tr '\t' '\n' | sed -n '/FORMAT/,$p' | tail -n +2  > exclude_tmp2.txt
fgrep -v -f exclude_tmp1.txt exclude_tmp2.txt > cluster_specific_maf_filters/exclude_tmp3.txt
cat cluster_specific_maf_filters/exclude_tmp3.txt <(awk '{print $1}' Pk.dups) lab_isolates.txt | sort | uniq > cluster_specific_maf_filters/hmmIBD/exclude_for_Mn.txt

awk '{print $2}' cluster_specific_maf_filters/Mn.bim | sed 's/:/\t/g' | awk '{print $1"\t"$2}' | uniq -u > cluster_specific_maf_filters/hmmIBD/Mn_grep_patterns.tsv

awk '{print $1}' cluster_specific_maf_filters/Pen.fam > exclude_tmp1.txt
bcftools view -h Consensus_SNPs_subset_no_MOI.vcf.gz | grep '#CHROM' | tr '\t' '\n' | sed -n '/FORMAT/,$p' | tail -n +2 > exclude_tmp2.txt
fgrep -v -f exclude_tmp1.txt exclude_tmp2.txt > cluster_specific_maf_filters/exclude_tmp3.txt
cat cluster_specific_maf_filters/exclude_tmp3.txt <(awk '{print $1}' Pk.dups) lab_isolates.txt | sort | uniq > cluster_specific_maf_filters/hmmIBD/exclude_for_Pen.txt

awk '{print $2}' cluster_specific_maf_filters/Pen.bim | sed 's/:/\t/g' | awk '{print $1"\t"$2}' | uniq -u > cluster_specific_maf_filters/hmmIBD/Pen_grep_patterns.tsv

rm -f exclude_tmp1.txt exclude_tmp2.txt cluster_specific_maf_filters/*tmp*
```


# Create genotype file for each

Rscript /g/data/pq84/malaria/Parasite_and_human_genetic_risk_factors_for_Pk_malaria/scripts/05_Analyses/Population_genetics/hmmIBD/genotype_clusters.R

```{r,eval=F}
# Packages
library(dplyr)
library(readr)
library(stringr)
library(data.table)

# Read in and wrangle VCF

create_IBD_genotype <- function(FILE_PATH){
    # Dynamically find the #CHROM header line number
    vcf_path <- "cluster_specific_maf_filters/hmmIBD/Consensus_SNPs_subset_no_MOI.vcf"
    vcf_lines <- readLines(vcf_path)
    header_line <- which(grepl("^#CHROM", vcf_lines))
    
    read_tsv(FILE_PATH, col_names = c("CHROM", "POS")) %>%
        mutate(POS = as.numeric(POS)) %>%
        left_join(
            read_table(vcf_path, skip = header_line - 1) %>%
                rename("CHROM" = `#CHROM`) %>%
                mutate(POS = as.numeric(POS)) 
        ) %>%
        mutate(CHROM = str_remove(CHROM, "ordered_PKNH_"),
               CHROM = str_remove(CHROM, "_v2")) %>%
        mutate_at(c(10:ncol(.)), ~str_remove(., ":.*")) %>% 
        mutate_at(c(10:ncol(.)), ~ifelse(. %like% "1/1" | . %like% "1/0" | . %like% "0/1" , "1", .)) %>% 
        mutate_at(c(10:ncol(.)), ~ifelse(. %like% "2/2" | . %like% "2/0" | . %like% "0/2" , "2", .)) %>% 
        mutate_at(c(10:ncol(.)), ~ifelse(. %like% "3/3" | . %like% "3/0" | . %like% "0/3" , "3", .)) %>% 
        mutate_at(c(10:ncol(.)), ~ifelse(. %like% "4/4" | . %like% "4/0" | . %like% "0/4" , "4", .)) %>% 
        mutate_at(c(10:ncol(.)), ~ifelse(. %like% "0/0" , "0", .)) %>% 
        mutate_at(c(10:ncol(.)), ~ifelse(. %like% "./." , "-1", .)) %>% 
        select(-c(3:9)) %>%
        arrange(CHROM, POS) %>%
        write_tsv(paste0(str_replace(FILE_PATH, "_grep_patterns.tsv", "_hmmIBD.tsv")))
}

list_of_files <- list.files(path = "cluster_specific_maf_filters/hmmIBD", pattern = "patterns.tsv", full.names = TRUE) 

for (i in list_of_files){
    create_IBD_genotype(paste0(i))
}
```

# Run hmmIBD

```{bash,eval=F}
cd cluster_specific_maf_filters/hmmIBD/
hmmIBD -i Mf_hmmIBD.tsv -b exclude_for_Mf.txt -o Mf
hmmIBD -i Mn_hmmIBD.tsv -b exclude_for_Mn.txt -o Mn
hmmIBD -i Pen_hmmIBD.tsv -b exclude_for_Pen.txt -o Pen
```

Rscript /g/data/pq84/malaria/pk_global_popgen/scripts/hmmIBD_stats.R 
Rscript /g/data/pq84/malaria/pk_global_popgen/scripts/hmmIBD_plots_cluster_spec.R 
Rscript /g/data/pq84/malaria/pk_global_popgen/scripts/hmmIBD_explore_clusters.R 

# Selection



# Variant selection - [rehh](https://cran.r-project.org/web/packages/rehh/vignettes/rehh.html#LoadData)

Use clonal_dups.txt to filter out clonal samples - messes with alogrithm.


Since the whole file is read in for rehh, even though you can only look at a single chr at a time, it is advisable to split large data sets into chromosome-specific files.
In addition, computing Rsb requires two data frames containing each the inES statistics of one population as obtained by the scan_hh() function - THUS - need to create separate VCF files for the different clusters that we want to compare.

Steps:
    - filter VCF using high quality GWAS type set of variants produced through plink (high_quality_set)
    - remove samples not in high quality set, dups_from_naming_error.txt, lab_isolates.txt, clonal_dups.txt
    - covert missing data to reference - otherwise they're filtered out by rehh
    - create a lsit of the chromosomes
    - get a list of samples to be included (-S) in the rehh analysis (Mn of Mf)
    - subset VCF into a VCF for individual chromosomes
    - subset to clusters

**Need to ammend the sn and xl line manually based on outputs of "cat"**


Pulling exclude samples from above and adding to clonal duplicates


```{bash,eval=F}
cd /g/data/pq84/malaria/pk_global_popgen/outputs/merged
mkdir -p selection/cluster_spec/Mf

awk '{print $2}' cluster_specific_maf_filters/Mf.bim | sed 's/:/\t/g' | awk '{print $1 "\t" $2}' > selection/cluster_spec/Mf/snp_filter.txt
cp cluster_specific_maf_filters/hmmIBD/exclude_for_Mf.txt selection/cluster_spec/Mf/exclude.txt
bcftools view --force-samples -R selection/cluster_spec/Mf/snp_filter.txt -S ^selection/cluster_spec/Mf/exclude.txt Consensus_SNPs_subset_no_MOI.vcf.gz | bcftools +missing2ref -o selection/cluster_spec/Mf/selection.vcf.gz
bcftools index selection/cluster_spec/Mf/selection.vcf.gz
bcftools view selection/cluster_spec/Mf/selection.vcf.gz | grep '##contig=<ID=ordered' | sed 's/##contig=<ID=//g' | sed 's/,.*//g' > selection/cluster_spec/chromosome.txt

for i in $(cat selection/cluster_spec/chromosome.txt)
    do 
        bcftools view -r $i -o selection/cluster_spec/Mf/${i}.vcf.gz selection/cluster_spec/Mf/selection.vcf.gz
    done 

mkdir -p selection/cluster_spec/Mn

awk '{print $2}' cluster_specific_maf_filters/Mn.bim | sed 's/:/\t/g' | awk '{print $1 "\t" $2}' > selection/cluster_spec/Mn/snp_filter.txt
cp cluster_specific_maf_filters/hmmIBD/exclude_for_Mn.txt selection/cluster_spec/Mn/exclude.txt
bcftools view --force-samples -R selection/cluster_spec/Mn/snp_filter.txt -S ^selection/cluster_spec/Mn/exclude.txt Consensus_SNPs_subset_no_MOI.vcf.gz | bcftools +missing2ref -o selection/cluster_spec/Mn/selection.vcf.gz
bcftools index selection/cluster_spec/Mn/selection.vcf.gz
bcftools view selection/cluster_spec/Mn/selection.vcf.gz | grep '##contig=<ID=ordered' | sed 's/##contig=<ID=//g' | sed 's/,.*//g' > selection/cluster_spec/chromosome.txt

for i in $(cat selection/cluster_spec/chromosome.txt)
    do 
        bcftools view -r $i -o selection/cluster_spec/Mn/${i}.vcf.gz selection/cluster_spec/Mn/selection.vcf.gz
    done 

mkdir -p selection/cluster_spec/Pen

awk '{print $2}' cluster_specific_maf_filters/Pen.bim | sed 's/:/\t/g' | awk '{print $1 "\t" $2}' > selection/cluster_spec/Pen/snp_filter.txt
cp cluster_specific_maf_filters/hmmIBD/exclude_for_Pen.txt selection/cluster_spec/Pen/exclude.txt
bcftools view --force-samples -R selection/cluster_spec/Pen/snp_filter.txt -S ^selection/cluster_spec/Pen/exclude.txt Consensus_SNPs_subset_no_MOI.vcf.gz \
    | bcftools +missing2ref -o selection/cluster_spec/Pen/selection.vcf.gz
bcftools index selection/cluster_spec/Pen/selection.vcf.gz
bcftools view selection/cluster_spec/Pen/selection.vcf.gz | grep '##contig=<ID=ordered' | sed 's/##contig=<ID=//g' | sed 's/,.*//g' > selection/cluster_spec/chromosome.txt

for i in $(cat selection/cluster_spec/chromosome.txt)
    do 
        bcftools view -r $i -o selection/cluster_spec/Pen/${i}.vcf.gz selection/cluster_spec/Pen/selection.vcf.gz
    done 
```


# Selection with rehh 

Rscript /g/data/pq84/malaria/Parasite_and_human_genetic_risk_factors_for_Pk_malaria/scripts/05_Analyses/Population_genetics/rehh.R


```{r,eval=F}
# Packages
library(rehh)
library(tidyverse)
library(data.table)
library(R.utils)

# WHOLE GENOME iHS

## need to read in chr files independently and build up an object resulting from the scan_hh outputs
### first create an empty df to be appended in the loop and a list of chr names
### the loop then creates the vcf file name (hap_file), and reads in the file to hh, which is then parsed into the scan_hh function to produce an object i that is appended to the growing wgscan object
### the resulting wgscan produces the same results as the scan_hh object above
### this can then be parsed into ihh2ihs

chromsomes_files <- c("ordered_PKNH_01_v2", "ordered_PKNH_02_v2", "ordered_PKNH_03_v2", "ordered_PKNH_04_v2", "ordered_PKNH_05_v2",
    "ordered_PKNH_06_v2", "ordered_PKNH_07_v2", "ordered_PKNH_08_v2", "ordered_PKNH_09_v2", "ordered_PKNH_10_v2", "ordered_PKNH_11_v2",
    "ordered_PKNH_12_v2", "ordered_PKNH_13_v2", "ordered_PKNH_14_v2")

wgscan <- tibble("CHR" = NA, "POSITION" = NA, "FREQ_A" = NA, "FREQ_D" = NA, "NHAPLO_A" = NA, "NHAPLO_D" = NA, "IHH_A" = NA, "IHH_D" = NA, "IES" = NA, "INES" = NA)

for(i in chromsomes_files) {
    # haplotype file name for each chromosome
    hap_file = paste(i, ".vcf.gz", sep = "") # filename pattern
    # create internal representation
    hh <- data2haplohh(hap_file = hap_file, 
                    chr.name = i, 
                    polarize_vcf = FALSE, # if the AA key is absent
                    min_perc_geno.mrk = 100, # discard markers genotyped on < 100% of haplotypes
                    min_maf = 0.05, # discard markers with a minor allele frequency of < 0.05)
                    vcf_reader = "data.table") # use this package
    scan <- scan_hh(hh) # perform scan on a single chromosome (calculate iHH/iES values)
    wgscan <- wgscan %>% # append the wgscan object every round with additional data
    rbind(scan)
}

wgs_ihs <- wgscan %>%
    na.omit() %>%
    ihh2ihs(., min_maf = 0.05, # default
            freqbin = 0.025 # default
            )

### Output is a list with two elements:
### ihs: a data frame with iHS and corresponding p-value piHS (p-value in a −log10 scale assuming the iHS are normally distributed under the neutral hypothesis)
### frequency.class: a data frame summarizing the derived allele frequency bins used for standardization and mean and standard deviation of the un-standardized values
wgs_ihs$ihs
wgs_ihs$frequency.class


# Scan for candidate regions 
## First update iHS table so that its compatible
wgs_ihs_2 <- wgs_ihs
wgs_ihs_2$ihs <- wgs_ihs_2$ihs %>%
    mutate(CHR = str_remove(CHR, "ordered_PKNH_")) %>%
    mutate(CHR = str_remove(CHR, "_v2")) %>%
    mutate(CHR = as.numeric(CHR)) %>%
    as.data.frame()

candidate_regions <- calc_candidate_regions(wgs_ihs_2,
                                 threshold = 4,
                                 pval = TRUE,
                                 window_size = 10000,
                                 overlap = 1000,
                                 min_n_extr_mrk = 3) %>%
                                 add_column(Stat = "iHS") 

## Save candidate regions for all stats
candidate_regions %>%
  write_tsv("candidate_regions_iHS.tsv")

selection_plots <- function(DATA, MODEL){
  plot_data <- DATA %>%
    na.omit() %>%
    mutate(CHR = str_remove(CHR, "ordered_PKNH_")) %>%
    mutate(CHR = str_remove(CHR, "_v2")) %>%
    mutate(CHR = as.numeric(CHR)) %>%
    arrange(CHR, POSITION) %>%
    mutate(POS = 1:nrow(.)) %>%
    mutate(CHR = as.factor(CHR))

  x_axis <- plot_data %>%
    group_by(CHR) %>%
    summarise(POS = median(POS)) %>%
    mutate(CHR = str_remove(CHR, "^0"))

  hline <- plot_data %>% 
    summarise(enframe(quantile(!!sym(MODEL), c(0.001, 0.5, 0.999)), "quantile", MODEL))

  selection_plot <- plot_data %>%
    ggplot(aes(x = POS, y = !!sym(MODEL), colour = CHR)) +
    geom_point() + 
    theme(legend.position = "none") +
    scale_colour_manual(values = c("#39568CFF", "#29AF7FFF", "#39568CFF", "#29AF7FFF", "#39568CFF", "#29AF7FFF", "#39568CFF", "#29AF7FFF", "#39568CFF", "#29AF7FFF", "#39568CFF", "#29AF7FFF", "#39568CFF", "#29AF7FFF")) +
    scale_x_continuous(breaks = x_axis$POS, labels = x_axis$CHR) +
    xlab("Position") +
    ylab(MODEL) +
    geom_hline(yintercept = c(as.numeric(hline[1,2]), as.numeric(hline[3,2])), linetype="dotted")

  ggsave(paste0(MODEL, ".png"), dpi = 300, width = 16, selection_plot)

  selection_plot <- plot_data %>%
      ggplot(aes(x = POS, y = LOGPVALUE, colour = CHR)) +
      geom_jitter() + 
      theme(legend.position = "none") +
      scale_x_continuous(breaks = x_axis$POS, labels = x_axis$CHR) +
      scale_colour_manual(values = c("#39568CFF", "#29AF7FFF", "#39568CFF", "#29AF7FFF", "#39568CFF", "#29AF7FFF", "#39568CFF", "#29AF7FFF", "#39568CFF", "#29AF7FFF", "#39568CFF", "#29AF7FFF", "#39568CFF", "#29AF7FFF")) +
      xlab("Chromosomes") +
      ylab(expression(MODEL ~ ~-log[10](italic(p))))

  ggsave(paste0(MODEL, "_pvalue.png"), dpi = 300, width = 16, selection_plot)
}

selection_plots(wgs_ihs$ihs, "IHS")
```


# Selection between Aceh and Peninsular Malaysia within Peninsular cluster - Rsb

```{R,eval=F}
library(tidyverse)
## Read in data
output_path = "/g/data/pq84/malaria/pk_global_popgen/outputs/merged/"
admix_clusters <- read_tsv(paste0(output_path, "outputs/admix_clusters.tsv"))
dups <- read.table(paste0(output_path, "Pk.dups"))
province_data <- read_tsv(paste0(output_path, "outputs/location_data.tsv")) %>%
	select(-Cluster) %>%
  select(Sample = ID, district, State, Country, Fws, enroldate)

metadata <- admix_clusters %>%
  select(Sample, Cluster) %>%
  filter(!Sample %in% dups$V1) %>%
  left_join(province_data, by = "Sample")

# Subset to 1 clone per cluster
clones <- read_tsv("/g/data/pq84/malaria/pk_global_popgen/outputs/merged/cluster_specific_maf_filters/hmmIBD/Pen_clones_IBD.txt")

## Build undirected graph where edges connect clones
clone_graph <- graph_from_data_frame(clones, directed = FALSE)

## Find connected components (clonal clusters)
clusters <- components(clone_graph)

## Turn into a dataframe
clonal_membership <- tibble(
    Sample = names(clusters$membership),
    CloneGroup = clusters$membership
    )

## Get rep with lowest missingness
miss <- read_table("Pk.imiss", col_names = T) %>%
	select(Sample = FID, F_MISS)

## Merge clone groups with missingness
clonal_with_miss <- clonal_membership %>%
  left_join(miss, by = "Sample")

## Select the sample with lowest missingness per clonal group
clonal_representatives <- clonal_with_miss %>%
  group_by(CloneGroup) %>%
  slice_min(F_MISS, with_ties = FALSE) %>%
  ungroup()

# Get the samples to exclude (i.e., all but the representative)
samples_to_exclude <- setdiff(clonal_with_miss$Sample, clonal_representatives$Sample)


metadata %>%
    filter(Cluster == "Peninsular" & State == "Aceh") %>%
    select(1) %>%
    filter(!Sample %in% samples_to_exclude) %>%
    write_tsv("selection/cluster_spec/Pen/aceh_samples.txt", col_names = FALSE)

metadata %>%
    filter(Cluster == "Peninsular" & State == "Peninsular") %>%
    select(1) %>%
    write_tsv("selection/cluster_spec/Pen/pen_not_aceh_samples.txt", col_names = FALSE)
```

```{bash}
mkdir -p selection/cluster_spec/Pen/ache

bcftools view --force-samples -R selection/cluster_spec/Pen/snp_filter.txt -S selection/cluster_spec/Pen/aceh_samples.txt Consensus_SNPs_subset_no_MOI.vcf.gz \
    | bcftools +missing2ref -o selection/cluster_spec/Pen/ache/selection.vcf.gz

bcftools index selection/cluster_spec/Pen/ache/selection.vcf.gz
bcftools view selection/cluster_spec/Pen/ache/selection.vcf.gz | grep '##contig=<ID=ordered' | sed 's/##contig=<ID=//g' | sed 's/,.*//g' > selection/cluster_spec/chromosome.txt

for i in $(cat selection/cluster_spec/chromosome.txt)
    do 
        bcftools view -r $i -o selection/cluster_spec/Pen/ache/${i}.vcf.gz selection/cluster_spec/Pen/ache/selection.vcf.gz
    done


mkdir -p selection/cluster_spec/Pen/not_ache

bcftools view --force-samples -R selection/cluster_spec/Pen/snp_filter.txt -S selection/cluster_spec/Pen/pen_not_aceh_samples.txt Consensus_SNPs_subset_no_MOI.vcf.gz \
    | bcftools +missing2ref -o selection/cluster_spec/Pen/not_ache/selection.vcf.gz

bcftools index selection/cluster_spec/Pen/not_ache/selection.vcf.gz
bcftools view selection/cluster_spec/Pen/not_ache/selection.vcf.gz | grep '##contig=<ID=ordered' | sed 's/##contig=<ID=//g' | sed 's/,.*//g' > selection/cluster_spec/chromosome.txt

for i in $(cat selection/cluster_spec/chromosome.txt)
    do 
        bcftools view -r $i -o selection/cluster_spec/Pen/not_ache/${i}.vcf.gz selection/cluster_spec/Pen/not_ache/selection.vcf.gz
    done
```


```{bash,eval=F}
# Create output directories
mkdir -p selection/cluster_spec/Pen/ache
mkdir -p selection/cluster_spec/Pen/not_ache

# Step 1: Fill missing genotypes for the full dataset (once)
bcftools +missing2ref Consensus_SNPs_subset_no_MOI.vcf.gz -Oz -o selection/cluster_spec/Pen/filled_all_samples.vcf.gz
bcftools index selection/cluster_spec/Pen/filled_all_samples.vcf.gz

# ---------------------
# ACEH subset
# ---------------------

# Step 2a: Subset to ACEH samples and desired SNPs
bcftools view --force-samples --min-ac=0 \
    -R selection/cluster_spec/Pen/snp_filter.txt \
    -S selection/cluster_spec/Pen/aceh_samples.txt \
    selection/cluster_spec/Pen/filled_all_samples.vcf.gz \
    -Oz -o selection/cluster_spec/Pen/ache/selection.vcf.gz

bcftools index selection/cluster_spec/Pen/ache/selection.vcf.gz

# Step 2b: Extract contig names and split per chromosome
bcftools view selection/cluster_spec/Pen/ache/selection.vcf.gz \
	| grep '##contig=<ID=ordered' \
	| sed 's/##contig=<ID=//g' | sed 's/,.*//g' \
	> selection/cluster_spec/chromosome.txt

for i in $(cat selection/cluster_spec/chromosome.txt)
    do 
        bcftools view -r $i -Oz -o selection/cluster_spec/Pen/ache/${i}.vcf.gz selection/cluster_spec/Pen/ache/selection.vcf.gz
    done

# ---------------------
# PENINSULAR (non-Aceh) subset
# ---------------------

# Step 3a: Subset to Peninsular (non-Aceh) samples and same SNPs
bcftools view --force-samples --min-ac=0 \
    -R selection/cluster_spec/Pen/snp_filter.txt \
    -S selection/cluster_spec/Pen/pen_not_aceh_samples.txt \
    selection/cluster_spec/Pen/filled_all_samples.vcf.gz \
    -Oz -o selection/cluster_spec/Pen/not_ache/selection.vcf.gz

bcftools index selection/cluster_spec/Pen/not_ache/selection.vcf.gz

# Step 3b: Use same contig list to split per chromosome
for i in $(cat selection/cluster_spec/chromosome.txt)
    do 
        bcftools view -r $i -Oz -o selection/cluster_spec/Pen/not_ache/${i}.vcf.gz selection/cluster_spec/Pen/not_ache/selection.vcf.gz
    done

cd selection/cluster_spec/Pen
```

```{R,eval=F}
# Packages
library(rehh)
library(tidyverse)
library(data.table)
library(R.utils)

# Comparing EHH between clusters

# CALCULATE Rsb  and XP-EHH

## pairwise population statistic 
## ratio of EHHS between populations

chromsomes_files <- c("ordered_PKNH_01_v2", "ordered_PKNH_02_v2", "ordered_PKNH_03_v2", "ordered_PKNH_04_v2", "ordered_PKNH_05_v2",
    "ordered_PKNH_06_v2", "ordered_PKNH_07_v2", "ordered_PKNH_08_v2", "ordered_PKNH_09_v2", "ordered_PKNH_10_v2", "ordered_PKNH_11_v2",
    "ordered_PKNH_12_v2", "ordered_PKNH_13_v2", "ordered_PKNH_14_v2")

### Ache
wgscan_ache <- tibble("CHR" = NA, "POSITION" = NA, "FREQ_A" = NA, "FREQ_D" = NA, "NHAPLO_A" = NA, "NHAPLO_D" = NA, "IHH_A" = NA, "IHH_D" = NA, "IES" = NA, "INES" = NA)

for(i in chromsomes_files) {
    # haplotype file name for each chromosome
    hap_file = file.path("ache", paste0(i, ".vcf.gz"))  # for Ache loop    # create internal representation
    hh <- data2haplohh(hap_file = hap_file, 
                    chr.name = i, 
                    polarize_vcf = FALSE, # if the AA key is absent
                    min_perc_geno.mrk = 50, # discard markers genotyped on < 100% of haplotypes
                    min_maf = 0.01, # discard markers with a minor allele frequency of < 0.01)
                    vcf_reader = "data.table") # use this package
    scan <- scan_hh(hh) # perform scan on a single chromosome (calculate iHH/iES values)
    wgscan_ache <- wgscan_ache %>% # append the wgscan object every round with additional data
    rbind(scan)
}

### Not Ache
wgscan_pen <- tibble("CHR" = NA, "POSITION" = NA, "FREQ_A" = NA, "FREQ_D" = NA, "NHAPLO_A" = NA, "NHAPLO_D" = NA, "IHH_A" = NA, "IHH_D" = NA, "IES" = NA, "INES" = NA)

for(i in chromsomes_files) {
    # haplotype file name for each chromosome
    hap_file = file.path("not_ache", paste0(i, ".vcf.gz"))  # for Pen loop    # create internal representation
    hh <- data2haplohh(hap_file = hap_file, 
                    chr.name = i, 
                    polarize_vcf = FALSE, # if the AA key is absent
                    min_perc_geno.mrk = 50, # discard markers genotyped on < 100% of haplotypes
                    min_maf = 0.01, # discard markers with a minor allele frequency of < 0.01)
                    vcf_reader = "data.table") # use this package
    scan <- scan_hh(hh) # perform scan on a single chromosome (calculate iHH/iES values)
    wgscan_pen <- wgscan_pen %>% # append the wgscan object every round with additional data
    rbind(scan)
}


# Compute test statistics

# Rsb
Rsb <- ines2rsb(
        scan_pop1 = wgscan_ache,
        scan_pop2 = wgscan_pen,
        popname1 = "Aceh",
        popname = "Peninsular"
    )

# XP-EHH
XP_EHH <- ies2xpehh(
        scan_pop1 = wgscan_ache,
        scan_pop2 = wgscan_pen,
        popname1 = "Aceh",
        popname = "Peninsular"
    )

tibble(
    Total_SNPs = nrow(wgscan_ache),
    Valid_IHH_A_Ache = sum(!is.na(wgscan_ache$IHH_A)),
    Valid_IHH_A_Pen = sum(!is.na(wgscan_pen$IHH_A)),
    Shared = sum(!is.na(Rsb$RSB_Aceh_Peninsular))
    )


# Scan for candidate regions 
candidate_regions_Rsb <- calc_candidate_regions(Rsb,
                                 threshold = 4,
                                 pval = TRUE,
                                 window_size = 1000,
                                 overlap = 1000,
                                 min_n_extr_mrk = 1) %>%
                                 add_column(Stat = "Rsb")

candidate_regions_XP_EHH <- calc_candidate_regions(XP_EHH,
                                 threshold = 4,
                                 pval = TRUE,
                                 window_size = 1000,
                                 overlap = 1000,
                                 min_n_extr_mrk = 1) %>%
                                 add_column(Stat = "XP_EHH")

# No candidate regions found at default threshold of 4, so get top 20 values instead
Rsb_top20 <- Rsb %>%
    filter(!is.na(RSB_Aceh_Peninsular)) %>%
    arrange(desc(abs(RSB_Aceh_Peninsular))) %>%
    slice(1:20) %>%
    add_column(Stat = "Rsb")

XP_EHH_top20 <- XP_EHH %>%
    filter(!is.na(XPEHH_Aceh_Peninsular)) %>%
    arrange(desc(abs(XPEHH_Aceh_Peninsular))) %>%
    slice(1:20) %>%
    add_column(Stat = "XP_EHH")

Rsb_top20 %>% 
    rename(Aceh_Peninsular = RSB_Aceh_Peninsular) %>%
    rbind(
        XP_EHH_top20 %>%
            rename(Aceh_Peninsular = XPEHH_Aceh_Peninsular)
        ) %>%
    write_tsv("top_hits_across_models.tsv")



selection_plots <- function(DATA, MODEL){
  plot_data <- DATA %>%
    na.omit() %>%
    mutate(CHR = str_remove(CHR, "ordered_PKNH_")) %>%
    mutate(CHR = str_remove(CHR, "_v2")) %>%
    mutate(CHR = as.numeric(CHR)) %>%
    arrange(CHR, POSITION) %>%
    mutate(POS = 1:nrow(.)) %>%
    mutate(CHR = as.factor(CHR))

  x_axis <- plot_data %>%
    group_by(CHR) %>%
    summarise(POS = median(POS)) %>%
    mutate(CHR = str_remove(CHR, "^0"))

  hline <- plot_data %>% 
    summarise(enframe(quantile(!!sym(MODEL), c(0.001, 0.5, 0.999)), "quantile", MODEL))

  selection_plot <- plot_data %>%
    ggplot(aes(x = POS, y = !!sym(MODEL), colour = CHR)) +
    geom_point() + 
    theme(legend.position = "none") +
    scale_colour_manual(values = c("#39568CFF", "#29AF7FFF", "#39568CFF", "#29AF7FFF", "#39568CFF", "#29AF7FFF", "#39568CFF", "#29AF7FFF", "#39568CFF", "#29AF7FFF", "#39568CFF", "#29AF7FFF", "#39568CFF", "#29AF7FFF")) +
    scale_x_continuous(breaks = x_axis$POS, labels = x_axis$CHR) +
    xlab("Position") +
    ylab(MODEL) +
    geom_hline(yintercept = c(as.numeric(hline[1,2]), as.numeric(hline[3,2])), linetype="dotted")

  ggsave(paste0(MODEL, ".png"), dpi = 300, width = 16, selection_plot)

  selection_plot <- plot_data %>%
      ggplot(aes(x = POS, y = LOGPVALUE, colour = CHR)) +
      geom_jitter() + 
      theme(legend.position = "none") +
      scale_x_continuous(breaks = x_axis$POS, labels = x_axis$CHR) +
      scale_colour_manual(values = c("#39568CFF", "#29AF7FFF", "#39568CFF", "#29AF7FFF", "#39568CFF", "#29AF7FFF", "#39568CFF", "#29AF7FFF", "#39568CFF", "#29AF7FFF", "#39568CFF", "#29AF7FFF", "#39568CFF", "#29AF7FFF")) +
      xlab("Chromosomes") +
      ylab(expression(MODEL ~ ~-log[10](italic(p))))

  ggsave(paste0(MODEL, "_pvalue.png"), dpi = 300, width = 16, selection_plot)
}

selection_plots(wgs_ihs$ihs, "IHS")
```


# Fst between Aceh and Peninsular Malaysia within Peninsular cluster

## MAF specific filters 

Use clonal_dups.txt to filter out clonal samples - messes with alogrithm.


First create a file that contains the clusters you want use for the Fst. You can ammend the Pk.fam file (or metadata), but the file needs to be structured in PLINKs weird FID-IID-Cluster format.

Steps:
    - Apply filter (as done previously)
    - Attach metadata to fam file (as done previously)
    - Define clusters using metadata
    - Calculate and plot Fst for a-priori clusters defiend through ADMIXTURE (sample_clusters.txt)

### State based clusters 

```{R,eval=F}
library(tidyverse)

read_tsv("selection/cluster_spec/Pen/aceh_samples.txt", col_names = FALSE) %>% 
    add_column(V2 = "Aceh") %>%
    rbind(
        read_tsv("selection/cluster_spec/Pen/pen_not_aceh_samples.txt", col_names = FALSE) %>%
        add_column(V2 = "Peninsular")
    ) %>% 
    mutate(X2 = X1) %>%
    select(X1, X2, V2) %>% 
    write_tsv("fst/sample_clusters.txt", col_names = FALSE)
```

```{bash,eval=F}
mkdir -p fst 

# Fst
cat selection/cluster_spec/Pen/pen_not_aceh_samples.txt selection/cluster_spec/Pen/aceh_samples.txt | grep -f - cluster_specific_maf_filters/Pen.fam > fst/include_samples.txt

plink --bfile cluster_specific_maf_filters/Pen \
  --within fst/sample_clusters.txt \
  --keep fst/include_samples.txt \
  --fst \
  --allow-no-sex \
  --out fst/Pk

# Freq per pop

cat selection/cluster_spec/Pen/pen_not_aceh_samples.txt | grep -f - cluster_specific_maf_filters/Pen.fam > fst/include_samples_peninsular.txt
cat selection/cluster_spec/Pen/aceh_samples.txt  | grep -f - cluster_specific_maf_filters/Pen.fam > fst/include_samples_aceh.txt

plink --bfile cluster_specific_maf_filters/Pen \
  --keep fst/include_samples_aceh.txt\
  --freq \
  --allow-no-sex \
  --out fst/Pk_aceh

plink --bfile cluster_specific_maf_filters/Pen \
  --keep fst/include_samples_peninsular.txt \
  --freq \
  --allow-no-sex \
  --out fst/Pk_peninsular

cd fst
```

Mean Fst estimate:
Weighted Fst estimate:

### Plot Fst 

Fst_plot.R

```{R,eval=F}
library(tidyverse)

Fst_matrix <- read_table("Pk.fst", col_names=T) %>%
    mutate(CHR = str_remove(SNP, "ordered_PKNH_")) %>%
    mutate(CHR = str_remove(CHR, "_v2.*"))

freq_aceh <- read_table("Pk_aceh.frq", col_names = TRUE) %>%
    select(CHR, SNP, MAF_aceh = MAF)

freq_peninsular <- read_table("Pk_peninsular.frq", col_names = TRUE) %>%
    select(CHR, SNP, MAF_peninsular = MAF)

Fst_with_freq <- Fst_matrix %>%
    left_join(freq_aceh, by = c("SNP")) %>%
    left_join(freq_peninsular, by = c("SNP")) %>%
    mutate(
        FST = as.numeric(FST),
        direction = case_when(
        is.na(MAF_aceh) | is.na(MAF_peninsular) ~ NA_character_,
        MAF_aceh > MAF_peninsular ~ "Higher in Aceh",
        MAF_aceh < MAF_peninsular ~ "Higher in Peninsular",
        MAF_aceh == MAF_peninsular ~ "Same"
        )
    ) %>%
    select(-CHR)

plot_data <- Fst_matrix %>%
    filter(FST != "nan") %>%
    filter(FST > 0) %>%
    mutate(FST = as.numeric(FST)) %>%
    arrange(CHR, POS) %>%
    add_column(ROW = 1:nrow(.)) 

x_axis <- plot_data %>%
    group_by(CHR) %>%
    summarise(ROW = median(ROW)) %>%
    mutate(CHR = str_remove(CHR, "^0"))

Fst_plot <- plot_data %>%
    ggplot(aes(x = ROW, y = FST, colour = CHR)) +
    geom_point() +
    theme(legend.position = "none", text = element_text(size = 25)) +
    ylab("Fst") +
    xlab("Chromosome")  +
    scale_color_viridis_d("Chr") +
    guides(color=guide_legend(nrow=1, byrow=TRUE)) +
    scale_x_continuous(breaks = x_axis$ROW, labels = x_axis$CHR)

ggsave("Fst_manhattan.png", dpi = 300, width = 20, Fst_plot)

Fst_matrix %>%
    mutate(FST = as.numeric(FST)) %>%
    summarise(mean_fst = mean(FST, na.rm = TRUE)) %>%
    write_tsv("Fst_mean.txt")

outliers <- Fst_with_freq %>%
  filter(as.numeric(FST) > 0.9)

write_tsv(outliers, "fst_outliers_gt_0.9.tsv")
```


| XP-EHH  | Fst     | Interpretation |
|---------|---------|----------------|
| ✅ High | ❌ Low  | Suggests a **recent/ongoing sweep** — haplotypes are different despite similar allele frequencies. Strong XP-EHH signature likely reflects **haplotype structure**, not just allele drift. |
| ✅ High | ✅ High | Suggests **differential selection or local adaptation**. Long haplotypes and big allele frequency shifts → likely **sweep in one population**. |
| ❌ Low  | ✅ High | Suggests **allele frequency divergence**, possibly from **drift or structure**, not selection. |
| ❌ Low  | ❌ Low  | Likely **neutral**. No strong evidence of selection or drift. |


