---
title: "tess3r"
output: pdf_document
date: "2024-06-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = here::here())
source(here::here("scripts/_setup.R"))
dir.create(file.path(DATA_PROC, "MSC", "spatial"), showWarnings = FALSE, recursive = TRUE)
```

```{r}
# packages
library(LEA)
library(here)
library(tess3r)
library(tidyverse)
library(maps)
library(Rcpp)
library(rworldmap)
library(malariaAtlas)
library(rnaturalearth)   # 2026 Phase-4 add: sf coastline for ggplot maps
library(sf)
library(eSDM)            # pts2poly_centroids (used L385)
library(gridExtra)       # grid.arrange
library(MuMIn)           # r.squaredGLMM (used at end)
library(spdep)           # set.ZeroPolicyOption / joincount.mc / nb2listw
library(car)             # Anova (used in regression chunks)
library(janitor)         # row_to_names (DEM read)
library(terra)           # rast / terra raster I/O (referenced by tess3r workflow)
```

# Import data

```{r}
genotype <- ped2lfmm(file.path(DATA_RAW, "tess3r", "clusters_combined_fixed.ped"))
tes_genotype <- read.table(genotype)

## replace missing calls with NA 
unique(as.vector(as.matrix(tes_genotype[-1]))) # need to get rid of the 9
tes_genotype_clean <- tes_genotype %>%
     mutate_all(~ifelse(. == 9, NA, .))

original <- read.table("data/raw/tess3r/clusters_combined.fam")
gis <- read.table("data/raw/tess3r/complete_gis_na_omit.fam") 
gis <- gis[order(match(gis$V1, original$V1)),]
gis_ids <- gis$V1                      # keep IDs: needed to anchor Q columns to clusters
gis <- gis %>%
     select(V7, V8) %>% 
     mutate(long = ifelse(V7 < V8, V8, V7)) %>%
     mutate(lat = ifelse(V8 > V7, V7, V8)) %>%
     select(long, lat) %>%
     as.matrix()

nrow(tes_genotype_clean) == nrow(gis)
```

# Visualise distribution of GIS coords - quick/dirty version

```{r}
#png("gis_coordinates.png")
plot(gis, pch = 19, cex = .5, 
     xlab = "Longitude (°E)", ylab = "Latitude (°N)")
maps::map(add = T, interior = F)
#dev.off()
```

# Estimating acestry coefficients

```{r}
# cache the tess3r fit (K=1:5 ~ 7 min) so re-runs are instant; delete to refit
.tess3_cache <- file.path(DATA_PROC, "tess3_obj_K1-5.rds")
if (file.exists(.tess3_cache)) {
  tess3.obj <- readRDS(.tess3_cache)
} else {
  tess3.obj <- tess3(X = tes_genotype_clean, coord = gis, K = 1:5,
                     method = "projected.ls", ploidy = 2, openMP.core.num = 10)
  saveRDS(tess3.obj, .tess3_cache)
}

# Visualise distribution of root mean-squared errors computed on subset of loci for cross validation - quick/dirty version
#png("K_cross-validation_clean.png")
plot(tess3.obj, pch = 19, col = "blue",
     xlab = "Number of ancestral populations",
     ylab = "Cross-validation score")
#dev.off()
```

# Ancestry matrix plots

```{r}
# barplot
q.matrix <- qmatrix(tess3.obj, K = 3)
# Anchor the palette to what each Q column actually IS (see pk_q_palette in _setup.R).
my.palette <- pk_q_palette(q.matrix, gis_ids, 10)
#q.matrix <- read_tsv("data/raw/tess3r/q_matrix_clean.tsv", col_names = FALSE) %>% as.matrix()

# STRUCTURE-like barplot for the Q-matrix  (base R: write png + svg directly)
for (dev_fn in list(function() png(file.path(FIG_DIR, "tess3r_barplot.png"), width = 8, height = 6, units = "in", res = 300),
                    function() grDevices::pdf(file.path(FIG_DIR, "tess3r_barplot.pdf"), width = 8, height = 6),
                    function() svglite::svglite(file.path(FIG_DIR, "tess3r_barplot.svg"), width = 8, height = 6))) {
  dev_fn()
  barplot(q.matrix, border = NA, space = 0,
          xlab = "Individuals", ylab = "Ancestry proportions",
          main = "Ancestry matrix",
          col.palette = my.palette) -> bp
  dev.off()
}

# map
#png("barplot_clean2_k_4.png")
plot(q.matrix, gis, method = "map.max", interpol = FieldsKrigModel(10),  
     main = "Ancestry coefficients",
     xlab = "Longitude", ylab = "Latitude", 
     resolution = c(300,300), cex = .4,
     col.palette = my.palette)
#dev.off()
```

```{r}
my.palette <- pk_q_palette(q.matrix, gis_ids, 20)

map.polygon <- getMap(resolution = "high")

pl <- ggtess3Q(q.matrix, gis, map.polygon = map.polygon, col.palette = my.palette)
pl +
  #geom_sf(data = malaysia) +
  geom_path(data = map.polygon, aes(x = long, y = lat, group = group)) +
  #scale_x_continuous(limits = c(100, 120)) +
  #scale_y_continuous(limits = c(-0, 10)) +
  xlim(100, 120) + 
  ylim(-0, 10) + 
  coord_equal() + 
  geom_point(data = as.data.frame(gis), aes(x = long, y = lat), size = 0.2) + 
  xlab("Longitute") +
  ylab("Latitude") + 
  theme_pk_map() 

malaysia_shp <- pk_malaysia_shp()   # _setup.R: cached; malariaAtlas geoserver now 404s
map.polygon <- getMap(resolution = "high") # for lines on final plot

pl <- ggtess3Q(q.matrix, gis, map.polygon = malaysia_shp, col.palette = my.palette)

tess_plot <- pl +
  geom_sf(col = "transparent") + 
  geom_path(data = map.polygon, aes(x = long, y = lat, group = group)) +
  #scale_x_continuous(limits = c(100, 120)) +
  #scale_y_continuous(limits = c(-0, 10)) +
  xlim(100, 120) + 
  ylim(-0, 10) + 
  coord_equal() + 
  geom_point(data = as.data.frame(gis), aes(x = long, y = lat), size = 1, alpha = 0.5, colour = "#FDE725FF") + 
  xlab("Longitute") +
  ylab("Latitude") + 
  theme_pk_map() +
  coord_cartesian(expand = F)

save_fig(tess_plot, "tess_map", width = 10, height = 7)
```

# Using admixture data

## Import data
```{r}
# Read in admixture data
admixture_table <- read.table("data/raw/tess3r/cleaned.3.Q") %>%
    cbind(
        read.table("data/raw/tess3r/cleaned.fam") %>%
            select(1) %>%
            rename(Sample = V1)
    ) 

# Read in GIS data
gis <- read.table("data/raw/tess3r/complete_gis_na_omit.fam") 

# Filter admixture by GIS - remove samples with missing GIS - add convert to matrix for  ggtess3Q
admix_ids <- admixture_table %>% filter(Sample %in% gis$V1) %>% pull(Sample)
admixture_matrix <- admixture_table %>%
  filter(Sample %in% gis$V1) %>%
  select(-Sample) %>%
  as.matrix()

# Order GIS by admixture
gis <- gis[order(match(gis$V1, admixture_table$Sample)),]

# Pivot GIS for input into  ggtess3Q
gis <- gis %>%
     select(V7, V8) %>% 
     mutate(long = ifelse(V7 < V8, V8, V7)) %>%
     mutate(lat = ifelse(V8 > V7, V7, V8)) %>%
     select(long, lat) %>%
     as.matrix()

nrow(admixture_matrix) == nrow(gis)
```

# Plot

```{r}
pl <- ggtess3Q(admixture_matrix, gis, map.polygon = malaysia_shp, col.palette = my.palette)

admix_plot <- pl +
  geom_sf(col = "transparent") + 
  geom_path(data = map.polygon, aes(x = long, y = lat, group = group)) +
  #scale_x_continuous(limits = c(100, 120)) +
  #scale_y_continuous(limits = c(-0, 10)) +
  xlim(100, 120) + 
  ylim(-0, 10) + 
  coord_equal() + 
  geom_point(data = as.data.frame(gis), aes(x = long, y = lat), size = 1, alpha = 0.5, colour = "#FDE725FF") + 
  xlab("Longitute") +
  ylab("Latitude") + 
  theme_pk_map() +
  coord_cartesian(expand = F)

```

```{r}
# How many Mn samples?
admixture_table_cluster <- read.table("data/raw/tess3r/cleaned.3.Q") %>%
    cbind(
        read.table("data/raw/tess3r/cleaned.fam") %>%
            select(1) %>%
            rename(Sample = V1)
    ) %>% 
    arrange(desc(V3), desc(V2), desc(V1)) %>% 
    add_column(Sample2 = as.factor(1:nrow(.))) %>%
    pivot_longer(1:3, 
        names_to = "Ancestry", 
        values_to = "Proportion") %>% 
  mutate(Cluster = ifelse(Ancestry == "V1", "Mn",
                          ifelse(Ancestry == "V2", "Mf", "Peninsular"))) 


admixture_table_cluster %>%
  group_by(Sample) %>%
  summarise(Proportion = max(Proportion)) %>% 
  left_join(
    admixture_table_cluster
  ) %>%
  filter(Cluster == "Mn")
```


# Plot only Sabah

```{R}
# Q-surface ramp, anchored to the ADMIXTURE component -> cluster correspondence rather
# than applied in name order (was a hard-coded spectral ramp before the restyle).
my.palette <- pk_q_palette(admixture_matrix, admix_ids, 10)



admixture_table <- read.table("data/raw/tess3r/cleaned.3.Q") %>%
    cbind(
        read.table("data/raw/tess3r/cleaned.fam") %>%
            dplyr::select(1) %>%
            rename(Sample = V1)
    ) 

# Read in GIS data
gis <- read.table("data/raw/tess3r/complete_gis_na_omit.fam") %>%
  filter(!grepl("ERR", V1)) %>%
  filter(!grepl("SRR", V1))

# Filter admixture by GIS - remove samples with missing GIS - add convert to matrix for  ggtess3Q
admixture_matrix <- admixture_table %>%
  filter(Sample %in% gis$V1) %>%
  dplyr::select(-Sample) %>%
  as.matrix()

# Order GIS by admixture
gis <- gis[order(match(gis$V1, admixture_table$Sample)),]

# Wrangle GIS for input into  ggtess3Q
gis <- gis %>%
     dplyr::select(V7, V8) %>% 
     mutate(long = ifelse(V7 < V8, V8, V7)) %>%
     mutate(lat = ifelse(V8 > V7, V7, V8)) %>%
     dplyr::select(long, lat) %>%
     as.matrix()

# Check matrices are the same length - same sample number
nrow(admixture_matrix) == nrow(gis)

# Create polygon
pl <- ggtess3Q(admixture_matrix, gis, map.polygon = malaysia_shp, col.palette = my.palette)

# Create dataframe with long/lat and cluster for geom_point
tmp_gis <- read.table("data/raw/tess3r/complete_gis_na_omit.fam") %>%
  filter(!grepl("ERR", V1)) %>%
  filter(!grepl("SRR", V1))

admixture_tmp <- admixture_table %>%
  filter(Sample %in% tmp_gis$V1) 

geom_point_data <- gis %>%
  cbind(
    admixture_tmp
  ) %>%
  left_join(
    admixture_table_cluster %>%
      group_by(Sample) %>%
      summarise(Proportion = max(Proportion)) %>% 
      left_join(
        admixture_table_cluster
      ) %>%
      dplyr::select(Sample, Cluster)
  ) %>%
  dplyr::select(long, lat, Cluster, Sample)

# Plot polygon
sabah_admix_plot <- pl +
  geom_sf(col = "transparent") + 
  geom_path(data = map.polygon, aes(x = long, y = lat, group = group)) +
  xlim(115, 120) + 
  ylim(4, 7.5) + 
  coord_equal() + 
  geom_point(data = geom_point_data, aes(x = long, y = lat, shape = Cluster), size = 1, alpha = 0.5, colour = "#161A1D") + 
  xlab("Longitute") +
  ylab("Latitude") + 
  theme_pk_map() +
  coord_cartesian(expand = F)

#ggsave("figures/sabah_admix_map.png", dpi = 300, sabah_admix_plot, width = 10)
```

# Plot spatial distribution of mixed population samples

```{r}
# Ensure 'malaysia' sf is defined before the first geom_sf() use (chunk-ordering fix).
if (!exists("malaysia")) {
  malaysia <- ne_states(country = c("malaysia", "indonesia"), returnclass = "sf") %>%
    mutate(name = ifelse(name != "Sabah" & name != "Sarawak", "Other", name))
}

# All mixes
gis_admix_df <- admixture_table %>%
  mutate(Ancestry = ifelse((V1 > .05 & V2 > .05) | (V1 > .05 & V3 > .05) | (V2 > .05 & V3 > .05), "Mixed", "Single")) %>%
  #filter(Ancestry == "Mixed") %>% na.omit()
  left_join(
    geom_point_data
  ) %>%
  na.omit()

# Plot polygon
sabah_admix_plot <- ggplot() +
  geom_sf(data = malaysia) +
  geom_point(data = gis_admix_df, aes(x = long, y = lat, colour = Ancestry, shape = Cluster), alpha = 0.75, position = position_dodge(0.1)) +
  xlim(115, 120) + 
  ylim(4, 7.5) + 
  ylab("Latitude") +
  xlab("Longitude") +
  labs(colour = "Ancestral populations (n)", shape = "Genomic cluster") +
  scale_colour_pk_ordinal(name = "Ancestral populations (n)") +
  theme_pk_map()

save_fig(sabah_admix_plot, "sabah_admix_distribution", width = 10, height = 7)

# With Peninsular
gis_admix_df <- gis_admix_df %>%
  mutate(Ancestry = ifelse(V3 > .025, "Mixed Peninsular", "Single")) %>% 
  filter(Ancestry == "Mixed Peninsular")

# Plot polygon
ggplot() +
  geom_sf(data = malaysia) +
  geom_point(data = gis_admix_df, aes(x = long, y = lat, colour = Ancestry, shape = Cluster), alpha = 0.75, position = position_dodge(0.5)) +
  xlim(115, 120) + 
  ylim(4, 7.5) + 
  ylab("Latitude") +
  xlab("Longitude") +
  labs(colour = "Ancestral populations (n)", shape = "Genomic cluster") +
  scale_colour_pk_ordinal(name = "Ancestral populations (n)") +
  theme_pk_map()
```





# Clusters vs environmental data 

### Check for spatial-autocorrelation 

Adapted from: https://rspatial.org/analysis/3-spauto.html & https://mgimond.github.io/Spatial/spatial-autocorrelation-in-r.html 

```{r,warning=F,message=F}
## Using spdep function 
polygon_data <- geom_point_data %>% 
  #mutate(Cluster = ifelse(Cluster == "Mf", 1, 2)) %>% 
  #rename(n = Cluster) %>%
  mutate(Cluster = as.factor(Cluster)) %>%
  pts2poly_centroids(0.0225, crs = 4326, agr = "constant")

set.ZeroPolicyOption(TRUE) # needed for zero.policy argument - we have 0 values because some locations have no contiguity
spatial_weights <- poly2nb(polygon_data, snap = 0.0225) %>% # snap value based on size of polygon/radius
    nb2listw(style="W", zero.policy = TRUE)

# Categorical dependent - using data produced above
joincount.test(polygon_data$Cluster, spatial_weights)
joincount.mc(polygon_data$Cluster, spatial_weights, nsim = 999, zero.policy = TRUE)

# p > 0.05 - accept the null - values randomly distributed in study area
```

## Map plots showing distribution of sampling and rasters

### Anopheles Layer 

```{r}
# Get malayis map data
malaysia <- ne_states(country = c("malaysia", "indonesia"), returnclass = "sf") %>%
  mutate(name = ifelse(name != "Sabah" & name != "Sarawak", "Other", name))

# ALC raster — fetched here if not already in env (script was originally co-run with 02)
if (!exists("anoph_raster_df")) anoph_raster_df <- pk_anoph_raster_df()  # _setup.R cache

# Plot Malaysia with raster layer and district points added
plot_1 <- ggplot() +
  geom_sf(data = malaysia, col = "transparent") +
  scale_x_continuous(limits = c(107, 120)) +
  scale_y_continuous(limits = c(-4.5, 7)) +
  geom_raster(data = anoph_raster_df, mapping = aes(x = x, y = y, fill = z)) +
  geom_point(data = geom_point_data, aes(x = long, y = lat, shape = Cluster)) +
  scale_fill_pk_continuous(name = "ALC layer") +
  scale_colour_pk_cluster() +
  ylab("Latitude") +
  xlab("Longitude") +
  theme_pk_map() +
  theme(legend.position = "none")

plot_2 <- ggplot() +
  geom_sf(data = malaysia, col = "transparent") +
  geom_raster(data = anoph_raster_df, mapping = aes(x = x, y = y, fill = z)) +
  geom_point(data = geom_point_data, aes(x = long, y = lat, shape = Cluster, colour = Cluster)) +
  xlim(115, 120) + 
  ylim(4, 7.5) + 
  scale_fill_pk_continuous(name = "ALC layer") +
  scale_colour_pk_cluster() +
  ylab("Latitude") +
  xlab("Longitude") +
  theme_pk_map()

plot_raster <- grid.arrange(plot_1, plot_2, ncol = 2)
save_fig(plot_raster, "ALC_raster_full_dataset", width = 10, height = 6)
```

## Cluster and ALC 

### Subset anopheles data

#### create 5x5 matrices around GIS centroids
#### Conversion for distance in km to long/lat
one degree of long/lat = 111km
one degree divided / 111km * 2.5km
1/111*2.5=0.0225  

```{r,warning=F,message=F}
# Calculate upper and lower range
sample_coords_limits <- geom_point_data %>% 
  rename(x_gps = long, y_gps = lat) %>%
  mutate(x_gps_upper = x_gps + 0.02252) %>% 
  mutate(x_gps_lower = x_gps - 0.02252) %>% 
  mutate(y_gps_upper = y_gps + 0.02252) %>% 
  mutate(y_gps_lower = y_gps - 0.02252) %>% 
  dplyr::select(3:ncol(.)) %>% 
  mutate_if(is.character, as.factor)
```

### Create function and nest in loop to get mean(5x5) anoph data for each sample

```{r,warning=F,message=F}
get_anoph_data <- function(SAMPLE){
  
  sample_coords_temp <- sample_coords_limits %>% 
    filter(Sample == SAMPLE)
  
  anoph_raster_df %>%
    filter(x > sample_coords_temp$x_gps_lower & x < sample_coords_temp$x_gps_upper) %>% 
    filter(y > sample_coords_temp$y_gps_lower & y < sample_coords_temp$y_gps_upper) %>%
    add_column(Sample = SAMPLE)
  
}

# Create empty df to store outputs 
anoph_sample_data <- anoph_raster_df %>% 
  add_column(Sample = 0) %>% 
  dplyr::slice(-(1:nrow(.)))

# Loop into df
for(i in levels(sample_coords_limits$Sample)){
    anoph_sample_data <- anoph_sample_data %>%
        rbind(get_anoph_data(paste0(i)))
}    

test_data <- anoph_sample_data %>%
  dplyr::select(Sample, z) %>%
  group_by(Sample) %>%
  summarise(anoph_density = mean(z)) %>%
    left_join(
      geom_point_data %>%
        dplyr::select(Sample, Cluster)
    ) %>%
  mutate_if(is.character, as.factor)

# alias used by later joins (variable was renamed mid-script in the original)
anoph_data <- test_data
```

```{r}
## Linear
glm(anoph_density ~ Cluster, data = test_data) %>% 
  summary()

glm(anoph_density ~ Cluster, data = test_data, family="Gamma") %>% 
  summary()

## Logistic Regression - https://www.geo.fu-berlin.de/en/v/soga-r/Basics-of-statistics/Logistic-Regression/Logistic-Regression-in-R---An-Example/index.html
regression_model <- glm(Cluster ~ anoph_density, data = test_data, family = "binomial")
regression_model_2 <- glm(Cluster ~ log10(anoph_density), data = test_data, family = "binomial")

summary(regression_model) 
summary(regression_model_2) 
summary(regression_model)$coefficients 

exp(coefficients(regression_model)[2]) # for every unit increase in altitude the odds ration increases by 1.001899
exp(confint.default(regression_model)[2, ]) # very low confidence

car::Anova(regression_model) %>%
  as.data.frame(row.names = NULL)

anova(regression_model, tes = "Chisq") # overall effect

test_data %>% 
  ggplot(aes(x = Cluster, y = anoph_density)) + 
  geom_boxplot()

r.squaredGLMM(regression_model) # pseudo R2
```


# Is there a relationship between cluster and elevation

## Export data to get DEM data
https://www.gpsvisualizer.com/elevation

```{r}
geom_point_data %>%
  dplyr::select(1:2) %>%
  write_tsv(file.path(DATA_PROC, "MSC", "gis_for_DEM.txt"), col_names = FALSE)
```

```{r}
test_data <- read.table("data/raw/MSC/DEM.txt", sep = "\t") %>%
  row_to_names(1) %>%
  dplyr::select(2:4) %>%
  rename(altitude = "altitude (m)") %>%
  cbind(
    geom_point_data # double check that the lat and long line up - rounding might cause some confusion
  ) %>%
  mutate(altitude = as.numeric(altitude)) %>%
  mutate(altitude = ifelse(altitude < 0.1, 0.1, altitude)) %>%# gamma models can't handle negative values
  mutate_if(is.character, as.factor)

shapiro.test(test_data$altitude)

test_data %>%
  ggplot(aes(x = altitude)) +
  geom_histogram()

## Linear
glm(altitude ~ Cluster, data = test_data) %>% 
  summary()

glm(altitude ~ Cluster, data = test_data, family="Gamma") %>% 
  summary()

## Logistic Regression - https://www.geo.fu-berlin.de/en/v/soga-r/Basics-of-statistics/Logistic-Regression/Logistic-Regression-in-R---An-Example/index.html
regression_model <- glm(Cluster ~ altitude, data = test_data, family = "binomial")
regression_model_2 <- glm(Cluster ~ log10(altitude), data = test_data, family = "binomial")

summary(regression_model) 
summary(regression_model_2) 
summary(regression_model)$coefficients 

exp(coefficients(regression_model)[2]) # for every unit increase in altitude the odds ration increases by 1.001899
exp(confint.default(regression_model)[2, ]) # very low confidence

car::Anova(regression_model) %>%
  as.data.frame(row.names = NULL)

anova(regression_model, tes = "Chisq") # overall effect

test_data %>% 
  ggplot(aes(x = Cluster, y = altitude)) + 
  geom_boxplot()

r.squaredGLMM(regression_model) # pseudo R2
```

# Add elevation to tess3r plot

```{R}
# Q-surface ramp, anchored to the ADMIXTURE component -> cluster correspondence rather
# than applied in name order (was a hard-coded spectral ramp before the restyle).
my.palette <- pk_q_palette(admixture_matrix, admix_ids, 10)

# Plot polygon
sabah_admix_plot <- pl +
  geom_sf(col = "transparent") + 
  geom_path(data = map.polygon, aes(x = long, y = lat, group = group)) +
  xlim(115, 120) + 
  ylim(4, 7.5) + 
  coord_equal() + 
  geom_point(data = test_data, aes(x = long, y = lat, shape = Cluster, size = altitude), alpha = 0.5, colour = "#161A1D") + 
  xlab("Longitute") +
  ylab("Latitude") + 
  theme_pk_map() +
  coord_cartesian(expand = F) +
  scale_size(name = "Elevation")

save_fig(sabah_admix_plot, "sabah_admix_map", width = 10, height = 7)
```


## Are elevation and ALC correlated? YES

```{r}
test_data <- test_data %>%
  left_join(
     anoph_sample_data %>% 
      dplyr::select(Sample, z) %>% 
      group_by(Sample) 
  )

regression_model_1 <- glm(z ~ altitude, data = test_data)
regression_model_2 <- glm(z ~ log10(altitude), data = test_data)

summary(regression_model_1) 
summary(regression_model_2) 
summary(regression_model)$coefficients 

exp(coefficients(regression_model)[2]) # for every unit increase in altitude the odds ration increases by 1.001899
exp(confint.default(regression_model)[2, ]) # very low confidence

car::Anova(regression_model) %>%
  as.data.frame(row.names = NULL)

anova(regression_model, tes = "Chisq") # overall effect

test_data %>% 
  ggplot(aes(x = Cluster, y = altitude)) + 
  geom_boxplot()

r.squaredGLMM(regression_model) # pseudo R2
```


**Cluster is associated with both elevation and ALC distribution, but these two variables are co linear. This suggests to me that the vector distribution is the more likely driver, as elevation would effect vector distribution not the other way around. Then, is vector distribution impacted by environmental parameters?**

# Tree cover 

## Wrangle sample_coords data to make it appropriate fromat for [CSIRO app](https://shiny.csiro.au/landscape-metrics/) data download

```{r,warning=F,message=F}
geom_point_data %>% 
  dplyr::select(-Cluster) %>% 
  rename(site_id = Sample, location_x = long, location_y = lat) %>%
  relocate(site_id) %>%
  write_csv(file.path(DATA_PROC, "MSC", "spatial", "sample_gps_coords_large.csv"))
```

## Read in data

```{r,warning=F,message=F}
test_data <- read_csv("data/raw/MSC/spatial/Class_metrics_samples_large.csv") %>%
  mutate_if(is.character, as.factor) %>%
  rename(Sample = plot_id) %>%
  left_join(geom_point_data) 
```


```{R,eval=F}
# Perimeter:area ratio
test_data <- test_data %>% 
  filter(metric == "para_mn" & class_name == "Trees" & !is.na(n)) %>%
  mutate_if(is.character, as.factor)

## Logistic Regression - https://www.geo.fu-berlin.de/en/v/soga-r/Basics-of-statistics/Logistic-Regression/Logistic-Regression-in-R---An-Example/index.html
regression_model <- glm(Cluster ~ value, data = test_data, family = "binomial")
regression_model_2 <- glm(Cluster ~ log10(value), data = test_data, family = "binomial")

summary(regression_model) 
summary(regression_model_2) 

# Landscape Percentage
test_data <-  read_csv("data/raw/MSC/spatial/Class_metrics_samples_large.csv") %>%
  mutate_if(is.character, as.factor) %>%
  rename(Sample = plot_id) %>%
  left_join(geom_point_data)  %>% 
  filter(metric == "pland" & class_name == "Trees" & !is.na(n)) %>%
  mutate_if(is.character, as.factor)

## Logistic Regression - https://www.geo.fu-berlin.de/en/v/soga-r/Basics-of-statistics/Logistic-Regression/Logistic-Regression-in-R---An-Example/index.html
regression_model <- glm(Cluster ~ value, data = test_data, family = "binomial")
regression_model_2 <- glm(Cluster ~ log10(value), data = test_data, family = "binomial")

summary(regression_model) 
summary(regression_model_2) 
```

# Mixed model

Tree cover, elevation & vector distribution

```{r}
test_data <- read_csv("data/raw/MSC/spatial/Class_metrics_samples_large.csv") %>%
  mutate_if(is.character, as.factor) %>%
  rename(Sample = plot_id) %>%
  left_join(geom_point_data)  %>% 
  filter(metric == "pland" & class_name == "Trees" & !is.na(n)) %>%
  mutate_if(is.character, as.factor) %>% 
  rename(percetange_trees = value) %>%
  dplyr::select(Sample, percetange_trees) %>%
  left_join(
    read_csv("data/raw/MSC/spatial/Class_metrics_samples_large.csv") %>%
      mutate_if(is.character, as.factor) %>%
      rename(Sample = plot_id) %>%
      left_join(geom_point_data)  %>% 
      filter(metric == "para_mn" & class_name == "Trees" & !is.na(n)) %>%
      mutate_if(is.character, as.factor) %>% 
      rename(para_trees = value) %>%
      dplyr::select(Sample, para_trees)
  ) %>%
  left_join(
    read.table("data/raw/MSC/DEM.txt", sep = "\t") %>%
      row_to_names(1) %>%
      dplyr::select(2:4) %>%
      rename(altitude = "altitude (m)") %>%
      cbind(
        geom_point_data # double check that the lat and long line up - rounding might cause some confusion
      ) %>%
      mutate(altitude = as.numeric(altitude)) %>%
      mutate(altitude = ifelse(altitude < 0.1, 0.1, altitude)) %>%# gamma models can't handle negative values
      mutate_if(is.character, as.factor) %>%
      dplyr::select(Sample, altitude, Cluster)
  ) %>%
  left_join(
    anoph_data %>%
      dplyr::select(1:2)
  )


regression_model <- glm(Cluster ~ para_trees + percetange_trees + altitude + anoph_density, data = test_data, family = "binomial")
regression_model_2 <- glm(Cluster ~ log10(para_trees) + log10(percetange_trees) + log10(altitude) + log10(anoph_density), data = test_data, family = "binomial")

summary(regression_model)
summary(regression_model_2)
```

# Map plot of clonal samples

```{r}
clonal_samples <- read_tsv("data/raw/hmmIBD/IBD_Clonal_transmission_miss5.tsv") %>%
   rbind(
    read_tsv("data/raw/hmmIBD/Mn_clonal_transmission.tsv") %>%
      slice(-1)
        ) %>%
  left_join(
    geom_point_data %>%
      mutate(Sample = str_remove(Sample, "_DK.*")) %>%
      mutate(Sample = str_replace(Sample, "_", "-")) %>%
      rename(sample = Sample) %>%
      rename_all(~paste0(., "1")) 
  ) %>%
  left_join(
    geom_point_data %>%
      mutate(Sample = str_remove(Sample, "_DK.*")) %>%
      mutate(Sample = str_replace(Sample, "_", "-")) %>%
      rename(sample = Sample) %>%
      rename_all(~paste0(., "2")) 
  ) %>%
  filter(clonal_cluster != "Peninsular") %>%
  filter(
    str_remove(sample1, "-") != str_remove(sample2, "-")
  ) 

clonal_samples <- clonal_samples %>%
  dplyr::select(sample1, sample2, lat1, lat2, long1, long2, enroldate1, enroldate2, clonal_cluster) %>%
  add_column(cluster1 = 1:nrow(.)) %>% 
  mutate(cluster1 = ifelse(cluster1 == 6 | cluster1 == 2, 1, cluster1)) %>%
  na.omit() %>%
  arrange(cluster1) %>%  
  mutate(cluster1 = ifelse(cluster1 == 3, 2, 
                           ifelse(cluster1 == 4, 3,
                                  ifelse(cluster1 == 7, 4,
                                         ifelse(cluster1 == 8, 5, cluster1))))) %>%
  mutate(cluster2 = cluster1) %>%
  filter(sample1 != "NKB2-9" & cluster1 != 2 & cluster1 != 3)

clonal_samples <- clonal_samples %>%
  dplyr::select(ends_with("1"), clonal_cluster) %>%
  rename_all(~str_remove(., "1")) %>%
  rbind(
    clonal_samples %>%
      dplyr::select(ends_with("2"), clonal_cluster) %>%
      rename_all(~str_remove(., "2")) 
  ) %>% 
  mutate(cluster = ifelse(cluster == 4, 2,
                          ifelse(cluster == 5, 3, 
                                 ifelse(cluster == 9, 4,
                                        ifelse(cluster == 10, 5, 
                                               ifelse(cluster == 11, 6, cluster)))))) %>%
  mutate(cluster = as.factor(cluster))

clonal_cluster_plot <- ggplot() +
  geom_sf(data = malaysia) +
  geom_point(data = clonal_samples, aes(x = long, y = lat, colour = cluster, group = cluster, shape = clonal_cluster)) +
  geom_line(data = clonal_samples, aes(x = long, y = lat, colour = cluster, group = cluster), colour = "black") +
  xlim(115, 120) + 
  ylim(4, 7.5) + 
  ylab("Latitude") +
  xlab("Longitude") +
  labs(colour = "Clonal cluster", shape = "Genomic cluster") +
  scale_colour_viridis_d() +
  theme_pk_map()
  #scale_colour_manual(values = c("#73D055FF","#39568CFF", "#29AF7FFF"))

save_fig(clonal_cluster_plot, "clonal_cluster_plot", width = 10, height = 7)
```


# ADMIXTURE vs Introgression

V2 is Mf

## linear?
```{r}
intro_admix <- read_tsv("data/raw/Introgression/mn_samples_window_counts.tsv") %>%
  rbind(
    read_tsv("data/raw/Introgression/mf_samples_window_counts.tsv")
  ) %>% 
  rename(introgression = n, Sample = SAMPLE) %>%
  left_join(
    admixture_table
  ) %>% 
  left_join(
    geom_point_data %>% 
      dplyr::select(Sample, Cluster)
  ) %>%
  mutate(Mixed = as.factor(ifelse((V1 > .05 & V2 > .05) | (V1 > .05 & V3 > .05) | (V2 > .05 & V3 > .05), "Mixed", "Single")))


intro_admix %>%
  filter(Cluster == "Mf") %>%
  ggplot(aes(x = introgression, y = V2*100)) +
  geom_point() +
  ylab("Percentage Mf") +
  xlab("Introgression") +
  geom_smooth(method='lm', formula= y~x)

intro_admix %>%
  filter(Cluster == "Mn") %>%
  ggplot(aes(x = introgression, y = V1*100)) +
  geom_point() +
  ylab("Percentage Mn") +
  xlab("Introgression") +
  geom_smooth(method='lm', formula= y~x)
```

## mixed vs not mixed

```{r}
intro_admix %>%
  filter(Cluster != "Peninsular") %>% 
  ggplot(aes(x = Mixed, y = introgression)) + 
  geom_boxplot() +
  geom_jitter(aes(colour = Cluster)) +
  scale_colour_pk_cluster() +
  ylab(Introgression)
```
# Investigate introgressed regions in admixed samples

```{r}
## Linear
# Mf
glm(V2 ~ introgression, data = intro_admix, family="Gamma") %>% 
  summary()

#Mn
glm(V1 ~ introgression, data = intro_admix, family="Gamma") %>% 
  summary()

## Logistic Regression - https://www.geo.fu-berlin.de/en/v/soga-r/Basics-of-statistics/Logistic-Regression/Logistic-Regression-in-R---An-Example/index.html
regression_model <- glm(Mixed ~ introgression, data = intro_admix, family = "binomial")
regression_model_2 <- glm(Mixed ~ log10(introgression), data = intro_admix, family = "binomial")

summary(regression_model) 
summary(regression_model_2) 
summary(regression_model)$coefficients 

exp(coefficients(regression_model)[2]) # for every unit increase in altitude the odds ration increases by 1.001899
exp(confint.default(regression_model)[2, ]) # very low confidence

car::Anova(regression_model) %>%
  as.data.frame(row.names = NULL)

anova(regression_model, tes = "Chisq") # overall effect

test_data %>% 
  ggplot(aes(x = Cluster, y = anoph_density)) + 
  geom_boxplot()

r.squaredGLMM(regression_model) # pseudo R2
```


# Top windows

```{r}
filtered_intro <- read_tsv("data/raw/Introgression/introgressed_windows_filtered.tsv")

filtered_intro %>% 
  group_by(Cluster, WINDOW) %>% 
  summarise(n = n()) %>%
  arrange(desc(n)) %>% 
  left_join(
    read_tsv("data/raw/Introgression/mf_windows.tsv")
  ) %>% View()
```

# Subset to samples that are highly admixed - do they have similar regions of introgression


```{r}
highly_mixed_samples <- intro_admix %>% 
  filter(V1 > .1 & V2 > .1)

read_tsv("data/raw/Introgression/introgressed_windows_filtered.tsv") %>% 
  filter(SAMPLE %in% highly_mixed_samples$Sample) %>% 
  group_by(Cluster, WINDOW) %>% 
  summarise(n = n()) %>%
  arrange(desc(n))
```

# Regression analysis - introgression and env variables

```{r,warning=F,message=F}
test_data <- read_csv("data/raw/MSC/spatial/Class_metrics_samples_large.csv") %>%
  mutate_if(is.character, as.factor) %>%
  rename(Sample = plot_id) %>%
  left_join(geom_point_data)  %>% 
  filter(metric == "pland" & class_name == "Trees" & !is.na(n)) %>%
  mutate_if(is.character, as.factor) %>% 
  rename(percetange_trees = value) %>%
  dplyr::select(Sample, percetange_trees) %>%
  left_join(
    read_csv("data/raw/MSC/spatial/Class_metrics_samples_large.csv") %>%
      mutate_if(is.character, as.factor) %>%
      rename(Sample = plot_id) %>%
      left_join(geom_point_data)  %>% 
      filter(metric == "para_mn" & class_name == "Trees" & !is.na(n)) %>%
      mutate_if(is.character, as.factor) %>% 
      rename(para_trees = value) %>%
      dplyr::select(Sample, para_trees)
  ) %>%
  left_join(
    read.table("data/raw/MSC/DEM.txt", sep = "\t") %>%
      row_to_names(1) %>%
      dplyr::select(2:4) %>%
      rename(altitude = "altitude (m)") %>%
      cbind(
        geom_point_data # double check that the lat and long line up - rounding might cause some confusion
      ) %>%
      mutate(altitude = as.numeric(altitude)) %>%
      mutate(altitude = ifelse(altitude < 0.1, 0.1, altitude)) %>%# gamma models can't handle negative values
      mutate_if(is.character, as.factor) %>%
      dplyr::select(Sample, altitude, Cluster)
  ) %>%
  left_join(
    anoph_data %>%
      dplyr::select(1:2)
  ) %>%
  left_join(
    read_tsv("data/raw/Introgression/mf_samples_window_counts.tsv") %>%
        rbind(
          read_tsv("data/raw/Introgression/mn_samples_window_counts.tsv") 
          ) %>% 
        mutate(intro_level = ntile(n, 3))  %>% 
        mutate(intro_level = as.factor(ifelse(intro_level == 3, "High", ifelse(intro_level == 2, "Medium", "Low")))) %>%
        rename(Sample = SAMPLE)
  )

```


```{r}
# Check if cluster needs to be a covariate
glm(intro_level ~ Cluster, data = test_data, family = binomial()) %>% 
  summary()

# Logistic Regression - ordered - https://stats.oarc.ucla.edu/r/dae/ordinal-logistic-regression/
landscape_model <- polr(intro_level ~ anoph_density, data = test_data, Hess = TRUE) 
landscape_model_2 <- polr(intro_level ~ anoph_density * Cluster, data = test_data, Hess = TRUE) 
landscape_model_3 <- polr(intro_level ~ anoph_density + Cluster, data = test_data, Hess = TRUE) 


summary(landscape_model) 
summary(landscape_model_2) 
summary(landscape_model_3) # best fit

coef_table <- polr(intro_level ~ anoph_density + Cluster, data = test_data, Hess = TRUE)  %>%
  summary() %>%
  coef()

confint.default(landscape_model_3) 

p_value <- pnorm(abs(coef_table[, "t value"]), lower.tail = FALSE) * 2
cbind(coef_table, "p_value" = p_value)

# Logistic Regression - ordered - https://stats.oarc.ucla.edu/r/dae/ordinal-logistic-regression/
landscape_model <- polr(intro_level ~ para_trees, data = test_data, Hess = TRUE) 
landscape_model_2 <- polr(intro_level ~ para_trees * Cluster, data = test_data, Hess = TRUE) 
landscape_model_3 <- polr(intro_level ~ para_trees + Cluster, data = test_data, Hess = TRUE) 


summary(landscape_model) 
summary(landscape_model_2) # best fit
summary(landscape_model_3) 

coef_table <- polr(intro_level ~ para_trees * Cluster, data = test_data, Hess = TRUE)   %>%
  summary() %>%
  coef()

confint.default(landscape_model_2) 

p_value <- pnorm(abs(coef_table[, "t value"]), lower.tail = FALSE) * 2
cbind(coef_table, "p_value" = p_value)


# Logistic Regression - ordered - https://stats.oarc.ucla.edu/r/dae/ordinal-logistic-regression/
landscape_model <- polr(intro_level ~ percetange_trees, data = test_data, Hess = TRUE) 
landscape_model_2 <- polr(intro_level ~ percetange_trees * Cluster, data = test_data, Hess = TRUE) 
landscape_model_3 <- polr(intro_level ~ percetange_trees + Cluster, data = test_data, Hess = TRUE) 


summary(landscape_model) 
summary(landscape_model_2) 
summary(landscape_model_3) # best fit

coef_table <- polr(intro_level ~ percetange_trees + Cluster, data = test_data, Hess = TRUE)    %>%
  summary() %>%
  coef()

confint.default(landscape_model_3) 

p_value <- pnorm(abs(coef_table[, "t value"]), lower.tail = FALSE) * 2
cbind(coef_table, "p_value" = p_value)
```

# Set up data for testing presence/absecnc of specific windows (individual columns for windows added to test data above)

```{r,warning=F,message=F}
test_data_windows <- read_tsv("data/raw/Introgression/introgressed_windows_filtered.tsv") %>% 
  dplyr::select(SAMPLE, WINDOW, Cluster) %>% 
  rename(Sample = SAMPLE) %>% 
  filter(Sample %in% test_data$Sample) %>%
  add_column(Introgression = "Yes") %>%
  pivot_wider(names_from = WINDOW, values_from = Introgression) %>%
  mutate_all(funs(replace_na(., "No"))) %>% 
  dplyr::select(-Cluster) %>%
  rename_at(vars(2:ncol(.)), funs(paste0("Window_", .))) %>%
  left_join(
    test_data
  ) %>% 
  mutate_if(is.character, as.factor) 

column_names <- test_data_windows %>% 
  select_at(vars(contains("Window"))) %>%
  colnames() %>% 
  as.data.frame() %>% 
  rename(Window = ".")
```


# Presence/absence of introgressed window & landscape metrics

## Mixed effects regression models - effect of proportions of landscape on presence/absence of windows

```{r}
library(MuMIn)
# Perimeter:area ratio

log_reg_output <- tibble("Estimate" = NA, "Std. Error" = NA, "z value" = NA, "Pr(>|z|)" = NA, "Window" = NA, "Variable" = NA)

for(i in column_names$Window){
  log_reg_output <- log_reg_output %>% 
    rbind(
      glm(get(i) ~ para_trees * Cluster, data = test_data_windows, family = "binomial") %>%
        summary() %>%
        coef() %>%
        as.data.frame() %>%
        rownames_to_column(var = "Variable") %>%
        add_column(Window = paste0(i))
    )
}

sig_hits_PA <- log_reg_output %>%
  filter(Variable == "para_trees" | Variable == "para_trees:ClusterMn") %>%
  filter(`Pr(>|z|)` < 0.05)


# Percentage tree

log_reg_output <- tibble("Estimate" = NA, "Std. Error" = NA, "z value" = NA, "Pr(>|z|)" = NA, "Window" = NA, "Variable" = NA)

for(i in column_names$Window){
  log_reg_output <- log_reg_output %>% 
    rbind(
      glm(get(i) ~ percetange_trees * Cluster, data = test_data_windows, family = "binomial") %>%
        summary() %>%
        coef() %>%
        as.data.frame() %>%
        rownames_to_column(var = "Variable") %>%
        add_column(Window = paste0(i))
    )
}

sig_hits_perc_tree <- log_reg_output %>%
  filter(Variable == "percetange_trees" | Variable == "percetange_trees:ClusterMn") %>%
  filter(`Pr(>|z|)` < 0.05)


# Anoph
log_reg_output <- tibble("Estimate" = NA, "Std. Error" = NA, "z value" = NA, "Pr(>|z|)" = NA, "Window" = NA, "Variable" = NA)

for(i in column_names$Window){
  log_reg_output <- log_reg_output %>% 
    rbind(
      glm(get(i) ~ anoph_density * Cluster, data = test_data_windows, family = "binomial") %>%
        summary() %>%
        coef() %>%
        as.data.frame() %>%
        rownames_to_column(var = "Variable") %>%
        add_column(Window = paste0(i))
    )
}

# Elevation
log_reg_output <- tibble("Estimate" = NA, "Std. Error" = NA, "z value" = NA, "Pr(>|z|)" = NA, "Window" = NA, "Variable" = NA)

for(i in column_names$Window){
  log_reg_output <- log_reg_output %>% 
    rbind(
      glm(get(i) ~ altitude * Cluster, data = test_data_windows, family = "binomial") %>%
        summary() %>%
        coef() %>%
        as.data.frame() %>%
        rownames_to_column(var = "Variable") %>%
        add_column(Window = paste0(i))
    )
}

sig_hits_elevation <- log_reg_output %>%
  filter(Variable == "altitude" | Variable == "altitude:ClusterMn") %>%
  filter(`Pr(>|z|)` < 0.05)


sig_hits_PA %>% 
  rbind(sig_hits_perc_tree) %>%
  rbind(sig_hits_anoph) %>% 
  rbind(sig_hits_elevation)
  unique() %>%
  arrange((Window)) %>%
  left_join(
    read_tsv("data/raw/Introgression/mf_windows.tsv") %>% 
      dplyr::select(WINDOW, CHROM, start, end) %>%
      rbind(
        read_tsv("data/raw/Introgression/mf_windows.tsv") %>% 
          dplyr::select(WINDOW, CHROM, start, end) 
      ) %>%
      rename(Window = WINDOW) %>%
      mutate(Window = paste0("Window_", Window))
  ) %>% 
  left_join(
    filtered_intro %>% 
      group_by(WINDOW) %>% 
      rename(Window = WINDOW) %>%
      mutate(Window = paste0("Window_", Window)) %>%
      summarise(count = n())
  ) %>% 
  arrange(desc(count)) %>% View()
```

# Top introgressed regions - for biology
```{r}
filtered_intro %>% 
  group_by(WINDOW) %>% 
  summarise(n = n()) %>%
    left_join(
    read_tsv("data/raw/Introgression/mf_windows.tsv") %>% 
      dplyr::select(WINDOW, CHROM, start, end) %>%
      rbind(
        read_tsv("data/raw/Introgression/mf_windows.tsv") %>% 
          dplyr::select(WINDOW, CHROM, start, end) 
      )
  ) %>%
  unique() %>%
  arrange(desc(n))
```

# When did introgression events occur?

```{r}
filtered_intro <- read_tsv("data/raw/Introgression/introgressed_windows_filtered.tsv")

 read_tsv("data/raw/Introgression/mn_samples_window_counts.tsv") %>%
  rbind(read_tsv("data/raw/Introgression/mf_samples_window_counts.tsv")) %>% 
  rename(introgression = n, Sample = SAMPLE) %>%
  left_join(
    admixture_table
  ) %>% 
  left_join(
    geom_point_data %>% 
      dplyr::select(Sample, Cluster)
  ) %>%
  mutate(Mixed = as.factor(ifelse((V1 > .05 & V2 > .05) | (V1 > .05 & V3 > .05) | (V2 > .05 & V3 > .05), "Mixed", "Single"))) %>% 
   left_join()
 
 clonal_samples
```

# When did admixture occur?