---
title: "Spatial-introgression_analyses"
output: pdf_document
date: "2023-09-11"
---

# Load libraries
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = here::here())
source(here::here("scripts/_setup.R"))
# malariaAtlas >=1.6.0 dropped as.MAP*; shims preserve the original column shapes
# (x,y,z for rasters; long,lat,name_* for shapes) so downstream code is unchanged.
as.MAPraster <- function(r) {
  df <- terra::as.data.frame(r, xy = TRUE, na.rm = TRUE)
  stats::setNames(df, c("x", "y", "z"))
}
as.MAPshp <- function(s) {
  s_sf <- suppressWarnings(sf::st_as_sf(s))
  coords <- as.data.frame(sf::st_coordinates(s_sf))
  feat_col <- tail(grep("^L", names(coords), value = TRUE), 1)
  attrs <- as.data.frame(sf::st_drop_geometry(s_sf))
  attrs$.id <- seq_len(nrow(attrs))
  coords$.id <- coords[[feat_col]]
  m <- merge(coords, attrs, by = ".id", sort = FALSE)
  m <- m[order(m$.id), ]
  out <- cbind(long = m$X, lat = m$Y, m[, setdiff(names(m), c("X","Y",grep("^L",names(m),value=TRUE),".id")), drop = FALSE])
  rownames(out) <- NULL
  out
}
library(tidyverse)
library(here)
library(janitor)
library(gridExtra)
library(malariaAtlas)
library(sp)
library(readxl)
library(MASS)
library(spdep)
library(eSDM)
library(car)
library(terra)
library(rnaturalearth)
```

# Description 
This Rmd describes four separate analyses that explore the relationship between spatial factors and introgression events in Plasmodium Knowles:
 - Degree of introgression & anoph density
 - Degree of introgression & landscape metrics 
 - Presence/absence of introgressed window & anoph density 
 - Presence/absence of introgressed window & landscape metrics


# Anopheles density analysis

# Degree of introgression & anoph density

## Import, export & wrangle data

### Download & wrangle anopheles data from malaria atlas - get district-level anopheles density 

```{r,warning=F,message=F}
# What rasters are available (malariaAtlas 1.6.0+: listData defunct, use listRaster)
listRaster()

# Get data
malaysia_shp <- getShp(country = "Malaysia", admin_level = c("admin0", "admin1", "admin2"))
anoph_raster <- getRaster(dataset_id = "Explorer__2016_Anopheles_leucosphyrus_complex", shp = malaysia_shp) # only available for 2016
anoph_raster <- getRaster(dataset_id = "Explorer__2016_Anopheles_leucosphyrus_complex") # needed to pull out data

# Convert to df
anoph_raster_df <- as.MAPraster(anoph_raster) %>% 
  unclass %>% 
  as.data.frame() 

malaysia_shp_df <- as.MAPshp(malaysia_shp) %>% 
  unclass %>% 
  as.data.frame() 

# Identify what districts (shape data) the predictions (raster file) appear in
district_contour <- malaysia_shp_df %>% 
  filter(name_2 == "Betong") %>%
  dplyr::select(long, lat, name_1, name_2)

district_subset <- point.in.polygon(pol.x = district_contour$long, pol.y = district_contour$lat, 
                 point.x = anoph_raster_df$x, point.y = anoph_raster_df$y) %>%
  as.data.frame() %>%
  rename(Betong = ".") %>% 
  cbind(anoph_raster_df) %>% 
  filter(Betong == 1)

# Create function to loop over districts
what_district <- function(DISTRICT){
  district_contour <- malaysia_shp_df %>% 
    filter(name_2 == DISTRICT) %>%
    dplyr::select(long, lat, name_1, name_2)

  district_subset <- point.in.polygon(pol.x = district_contour$long, pol.y = district_contour$lat, 
                  point.x = anoph_raster_df$x, point.y = anoph_raster_df$y) %>%
    as.data.frame() %>%
    rename(District = ".") %>% 
    mutate(District = ifelse(District == 1, DISTRICT, .$District)) %>%
    cbind(anoph_raster_df) %>% 
    filter(District == DISTRICT)
  
  return(district_subset)
}

# Get district names
district_names <- malaysia_shp_df %>% 
  dplyr::select(name_2) %>% 
  na.omit() %>% 
  unique() %>% 
  mutate(name_2 = as.factor(name_2))

# Create empty df using raster data (same shape as input data)
district_level_vector_data <- anoph_raster_df %>% 
  add_column(District = 0) %>% 
  slice(-(1:nrow(.)))

# Loop into df
for(i in levels(district_names$name_2)){
    district_level_vector_data <- district_level_vector_data %>%
        rbind(what_district(paste0(i)))
}    

district_level_vector_data_summary <- district_level_vector_data %>% 
  group_by(District) %>% 
  summarise(mean = mean(z), median = median(z), min = min(z), max = max(z)) 

#write_tsv(district_level_vector_data_summary, "/Users/jacobwestaway/Desktop/Menzies/Knowlesi/Pipeline/Pk_Pipeline/data/MSC/district_level_vector_data.tsv")
district_level_vector_data %>% 
  dplyr::select(x, y)  
  #write_csv("/Users/jacobwestaway/Desktop/Menzies/Knowlesi/Pipeline/Pk_Pipeline/data/MSC/district_level_vector_data_coords.csv")
```


### Get GPS coords for all samples

```{r,warning=F,message=F}
#sample_coords <- read_xlsx("data/raw/metadata/PK_Sabah_Sample_naming_indexes.xlsx") %>% 
#  dplyr::select(sampleid, group, subjectid) %>%
#  mutate(subjectid = ifelse(str_length(subjectid) == 1, paste0("00", subjectid), # if subjectid length = 1 paste 00
#                     ifelse(str_length(subjectid) == 2, paste0("0", subjectid), # if not, then if subjectid length = 2 paste 0
#                     .$subjectid))) %>% # if not, keep value the same
#    mutate(studycode = paste0(group, subjectid)) %>%
#  left_join(
#      read_xlsx("data/raw/metadata/full_datasets/A_CC_master_19.6.16.xlsx") %>%
#        select(code, x_gpsbackup, y_gpsbackup) %>%
#        rename(studycode = code)
#  ) %>% 
#  filter(!is.na(x_gpsbackup)) %>%
#    arrange(sampleid)

# Updated coords from Matt

sample_coords <- read_csv("data/raw/MSC/spatial/pk_sample_coords_updated.csv") %>% 
  mutate(x_gpsbackup = ifelse(is.na(x_gpsbackup), .$x_long_final, .$x_gpsbackup)) %>% 
  mutate(y_gpsbackup = ifelse(is.na(y_gpsbackup), .$y_lat_final, .$y_gpsbackup)) %>% 
  dplyr::select(sampleid, code, x_gpsbackup, y_gpsbackup) %>% 
  rename(studycode = code) %>% 
  na.omit()
```


### Subset anopheles data

#### create 5x5 matrices around points
#### Conversion for distance in km to long/lat
one degree of long/lat = 111km
one degree divided / 111km * 2.5km
1/111*2.5=0.0225  

```{r,warning=F,message=F}
# Calculate upper and lower range
sample_coords_limits <- sample_coords %>% 
  rename(x_gps = x_gpsbackup, y_gps = y_gpsbackup) %>%
  mutate(x_gps_upper = x_gps + 0.02252) %>% 
  mutate(x_gps_lower = x_gps - 0.02252) %>% 
  mutate(y_gps_upper = y_gps + 0.02252) %>% 
  mutate(y_gps_lower = y_gps - 0.02252) %>% 
  dplyr::select(1, 5:ncol(.)) %>% 
  mutate_if(is.character, as.factor)
```


### Combine anopheles and introgression data

#### Create function and nest in loop to get anoph data for each sample

```{r,warning=F,message=F}
get_anoph_data <- function(SAMPLE){
  
  sample_coords_temp <- sample_coords_limits %>% 
    filter(sampleid == SAMPLE)
  
  anoph_raster_df %>%
    filter(x > sample_coords_temp$x_gps_lower & x < sample_coords_temp$x_gps_upper) %>% 
    filter(y > sample_coords_temp$y_gps_lower & y < sample_coords_temp$y_gps_upper) %>%
    add_column(sampleid = SAMPLE)
}

# Create empty df to store outputs 
anoph_sample_data <- anoph_raster_df %>% 
  add_column(sampleid = 0) %>% 
  dplyr::slice(-(1:nrow(.)))

# Loop into df
for(i in levels(sample_coords_limits$sampleid)){
    anoph_sample_data <- anoph_sample_data %>%
        rbind(get_anoph_data(paste0(i)))
}    
```

#### Join data

```{r,warning=F,message=F}
# Read in data
anoph_degree_of_intro <- anoph_sample_data %>% 
  dplyr::select(sampleid, z) %>% 
  group_by(sampleid) %>% 
  summarise(anoph_density = mean(z)) %>%
  left_join(
    read_tsv("data/raw/MSC/spatial/intro_per_sample_summary.tsv") %>%
      rename(sampleid = SAMPLE) %>% 
      mutate(sampleid = str_remove(sampleid, "_DK.*"))
  ) %>% 
  na.omit() %>%
  mutate_if(is.character, as.factor) %>%
  mutate(intro_level = fct_relevel(intro_level, c("Low","Medium","High")))
```


### Check for spatial-autocorrelation 

Adapted from: https://rspatial.org/analysis/3-spauto.html & https://mgimond.github.io/Spatial/spatial-autocorrelation-in-r.html 

```{r,warning=F,message=F}
# For continuous dependent

## Method 1
## Convert points to polygon
polygon_data <- anoph_degree_of_intro %>% 
  left_join(sample_coords) %>% 
  dplyr::select(x_gpsbackup, y_gpsbackup, sampleid, intro_level, n) %>% 
  pts2poly_centroids(0.0225, crs = 4326, agr = "constant") %>% # create polygons based on the 2.5km radius 
  vect() # convert to SpatVector object

## Check for intersecting polygons
plot(polygon_data)

## Compute Moran's I and do significance test
## Create spatial weights matrix
spatial_weights <- terra::adjacent(polygon_data, "intersect", pairs = FALSE) # intersects excludes intersecting polygons

## Calculate autocorrelation
ac_value <- autocor(polygon_data$n, spatial_weights, "moran") # continuous

## Test significance
mc_sim <- sapply(1:99, function(i) {
    autocor(sample(polygon_data$n), spatial_weights, "moran")
})

# Check simulated distribution of expected values against observed Moran's I - does the the observed value align with the distribution, which one would expect if the values were not spatially auto-correlated 
hist(mc_sim)
abline(v=ac_value, col="red")
sum(mc_sim >= ac_value) / 100

## Method 2
## Using spdep function 
polygon_data <- anoph_degree_of_intro %>% 
  left_join(sample_coords) %>% 
  dplyr::select(x_gpsbackup, y_gpsbackup, sampleid, intro_level, n) %>% 
  pts2poly_centroids(0.0225, crs = 4326, agr = "constant") 

set.ZeroPolicyOption(TRUE) # needed for zero.policy argument - we have 0 values because some locations have no contiguity
spatial_weights <- poly2nb(polygon_data, snap = 0.0225) %>% # snap value based on size of polygon/radius
    nb2listw(style="W", zero.policy = TRUE)

moran.test(polygon_data$n, spatial_weights)
moran.mc(polygon_data$n, spatial_weights, nsim = 999)

# Categorical dependent - using data produced above
joincount.test(polygon_data$intro_level, spatial_weights)
joincount.mc(polygon_data$intro_level, spatial_weights, nsim = 999, zero.policy = TRUE)

# p > 0.05 - accept the null - values randomly distributed in study area
```



### Check for relationships between Cluster and env variables, and introgression


#### Cluster and degree of introgression

```{r}
## Linear
glm(n ~ Cluster, data = anoph_degree_of_intro) %>% summary()

## Binomial
glm(intro_level ~ Cluster, data = anoph_degree_of_intro, family = binomial()) %>% 
  summary()

##  Ordinal
coef_table <- polr(intro_level ~ Cluster, data = anoph_degree_of_intro, Hess = TRUE)  %>%
  summary() %>%
  coef

p_value <- pnorm(abs(coef_table[, "t value"]), lower.tail = FALSE) * 2
cbind(coef_table, "p_value" = p_value)
```


#### Cluster and anoph density

```{r}
## Linear
glm(anoph_density ~ Cluster, data = anoph_degree_of_intro) %>% 
  summary()

glm(anoph_density ~ Cluster, data = anoph_degree_of_intro, family="Gamma") %>% 
  summary()
```

# Analyses
## Mixed effects regression models - effect of anoph density on degree of introgression 

```{r,warning=F,message=F}
# Logistic Regression - ordered - https://stats.oarc.ucla.edu/r/dae/ordinal-logistic-regression/
landscape_model <- polr(intro_level ~ anoph_density, data = anoph_degree_of_intro, Hess = TRUE) 
landscape_model_2 <- polr(intro_level ~ anoph_density * Cluster, data = anoph_degree_of_intro, Hess = TRUE) 
landscape_model_3 <- polr(intro_level ~ anoph_density + Cluster, data = anoph_degree_of_intro, Hess = TRUE) 


summary(landscape_model) 
summary(landscape_model_2) # model 2 is best fit but NOT significant
summary(landscape_model_3)

coef_table <- polr(intro_level ~ anoph_density * Cluster, data = anoph_degree_of_intro, Hess = TRUE) %>%
  summary() %>%
  coef()

confint.default(landscape_model_2) # significant association for interaction - anoph in Mn relative to Mf, but the same size is too small

p_value <- pnorm(abs(coef_table[, "t value"]), lower.tail = FALSE) * 2
cbind(coef_table, "p_value" = p_value)

anoph_degree_of_intro %>% 
  filter(Cluster == "Mn") %>%
  group_by(intro_level) %>%
  summarise(n()) # showing small sample size
```


###########################################################################################################################################################################

## Get major windows data and combine with anoph data - for downstream analysis

```{r,warning=F,message=F}
major_windows <- read_tsv("data/raw/MSC/spatial/Mf_major_windows_Sabah_samples.tsv") %>% 
  add_column(Cluster = "Mf") %>% 
  rbind(
    read_tsv("data/raw/MSC/spatial/Mf_major_windows_Sarawak_samples.tsv") %>%
      add_column(Cluster = "Mf") 
    ) %>%
  rbind(
    read_tsv("data/raw/MSC/spatial/Mn_major_windows_Sabah_samples.tsv") %>%
    add_column(Cluster = "Mn")
  ) %>% 
  rbind(
    read_tsv("data/raw/MSC/spatial/Mn_major_windows_Sarawak_samples.tsv") %>%
    add_column(Cluster = "Mn")
  ) %>%
  dplyr::select(WINDOW, Cluster) %>%
  unique()

anoph_sample_data <- anoph_sample_data %>% 
  dplyr::select(sampleid, z) %>% 
  group_by(sampleid) %>% 
  summarise(anoph_density = mean(z)) 

anoph_major_windows_summary <- read_tsv("data/raw/MSC/spatial/introgressed_windows_filtered.tsv") %>% 
  #filter(WINDOW %in% major_windows$WINDOW & Cluster %in% major_windows$Cluster) %>% 
  dplyr::select(SAMPLE, WINDOW, Cluster) %>% 
  rename(sampleid = SAMPLE) %>% 
  mutate(sampleid = str_remove(sampleid, "_DK.*")) %>% 
  filter(sampleid %in% anoph_sample_data$sampleid) %>%
  add_column(Introgression = "Yes") %>%
  pivot_wider(names_from = WINDOW, values_from = Introgression) %>%
  mutate_all(funs(replace_na(., "No"))) %>% 
  pivot_longer(3:ncol(.), names_to = "Window", values_to = "Introgression") %>% 
  filter(Window %in% major_windows$WINDOW & Cluster %in% major_windows$Cluster) %>% 
  pivot_wider(names_from = Window, values_from = Introgression) %>% 
  left_join(
    anoph_sample_data
  ) %>% 
  rename_at(vars(3:24), funs(paste0("Window_", .))) %>%
  mutate_if(is.character, as.factor)

column_names <- anoph_major_windows_summary %>% 
  select_at(vars(contains("Window"))) %>%
  colnames() %>% 
  as.data.frame() %>% 
  rename(Window = ".")
```


###########################################################################################################################################################################



# Landscape analysis

# Degree of introgression & landscape metrics

## Wrangle sample_coords data to make it appropriate fromat for [CSIRO app](https://shiny.csiro.au/landscape-metrics/) data download

```{r,warning=F,message=F}
sample_coords %>% 
  dplyr::select(sampleid,x_gpsbackup, y_gpsbackup) %>% 
  rename(site_id = sampleid, location_x = x_gpsbackup, location_y = y_gpsbackup) 
  #write_csv("data/raw/MSC/spatial/sample_gps_coords.csv")
```

## Read in data

```{r,warning=F,message=F}
landscape_sample_data <- read_csv("data/raw/MSC/spatial/Class_metrics_samples.csv") %>%
  mutate_if(is.character, as.factor) %>%
  rename(sampleid = plot_id) %>%
  left_join(anoph_degree_of_intro) %>%
  left_join(anoph_major_windows_summary) 
```

### Check for relationships between Cluster and env variables

#### Cluster and landscape

```{r}
# Perimeter:area ratio
landscape_sample_data_analysis <- landscape_sample_data %>% 
  filter(metric == "para_mn" & class_name == "Trees" & !is.na(n))

## Linear
glm(value ~ Cluster, data = landscape_sample_data_analysis) %>% 
  summary()

glm(value ~ Cluster, data = landscape_sample_data_analysis, family="Gamma") %>% 
  summary()

# Landscape Percentage
landscape_sample_data_analysis <- landscape_sample_data %>% 
  filter(metric == "pland" & class_name == "Trees" & !is.na(n))

glm(value ~ Cluster, data = landscape_sample_data_analysis) %>% 
  summary()

glm(value ~ Cluster, data = landscape_sample_data_analysis, family="Gamma") %>% 
  summary()
```


## Mixed effects regression models - effect of landscape proportions on degree of introgression 

```{r,warning=F,message=F}
# Perimeter:area ratio
landscape_sample_data_analysis <- landscape_sample_data %>% 
  filter(metric == "para_mn" & class_name == "Trees" & !is.na(n))

## Logistic Regression - ordered - https://stats.oarc.ucla.edu/r/dae/ordinal-logistic-regression/
library(MASS)

landscape_model <- polr(intro_level ~ value, data = landscape_sample_data_analysis, Hess = TRUE) 
landscape_model_2 <- polr(intro_level ~ value * Cluster, data = landscape_sample_data_analysis, Hess = TRUE) 
landscape_model_3 <- polr(intro_level ~ value + Cluster, data = landscape_sample_data_analysis, Hess = TRUE) 

summary(landscape_model) 
summary(landscape_model_2) 
summary(landscape_model_3) # model 3 is best fit

confint.default(landscape_model_2) # p-a ratio has no significant association with degree of intro

coef_table <- polr(intro_level ~ value + Cluster, data = landscape_sample_data_analysis, Hess = TRUE) %>%
  summary() %>%
  coef()

p_value <- pnorm(abs(coef_table[, "t value"]), lower.tail = FALSE) * 2
cbind(coef_table, "p_value" = p_value) # further validation
###################################################################################################

# Landscape Percentage
landscape_sample_data_analysis <- landscape_sample_data %>% 
  filter(metric == "pland" & class_name == "Trees" & !is.na(n))

## Logistic Regression - ordered - https://stats.oarc.ucla.edu/r/dae/ordinal-logistic-regression/
landscape_model <- polr(intro_level ~ value, data = landscape_sample_data_analysis, Hess = TRUE) 
landscape_model_2 <- polr(intro_level ~ value * Cluster, data = landscape_sample_data_analysis, Hess = TRUE) 
landscape_model_3 <- polr(intro_level ~ value + Cluster, data = landscape_sample_data_analysis, Hess = TRUE) 

summary(landscape_model) 
summary(landscape_model_2) 
summary(landscape_model_3) # model 3 is best fit

confint.default(landscape_model_2) # p-a ratio has no significant association with degree of intro

coef_table <- polr(intro_level ~ value + Cluster, data = landscape_sample_data_analysis, Hess = TRUE) %>%
  summary() %>%
  coef()

p_value <- pnorm(abs(coef_table[, "t value"]), lower.tail = FALSE) * 2
cbind(coef_table, "p_value" = p_value) # further validation
```



###########################################################################################################################################################################



# Presence/absence of introgressed window & landscape metrics

## Mixed effects regression models - effect of proportions of landscape on presence/absence of windows

```{r}
library(MuMIn)
# Perimeter:area ratio
landscape_sample_data_analysis <- landscape_sample_data %>% 
  filter(metric == "para_mn" & class_name == "Trees" & !is.na(n))

log_reg_output <- tibble("Estimate" = NA, "Std. Error" = NA, "z value" = NA, "Pr(>|z|)" = NA, "Window" = NA, "Variable" = NA)

for(i in levels(column_names$windows)){
  log_reg_output <- log_reg_output %>% 
    rbind(
      glm(get(i) ~ value * Cluster, data = landscape_sample_data_analysis, family = "binomial") %>%
        summary() %>%
        coef() %>%
        as.data.frame() %>%
        rownames_to_column(var = "Variable") %>%
        add_column(Window = paste0(i))
    )
  }

pa_ratio_windows <- log_reg_output %>% 
  na.omit() %>%
  rename(P = "Pr(>|z|)") %>%
  arrange(P) %>%
  filter(Variable == "value")

# Subset to significant (or close to) windows to reassess
landscape_model <- glm(Window_859 ~ value * Cluster, data = landscape_sample_data_analysis, family = "binomial") 
landscape_model_2 <- glm(Window_859 ~ value, data = landscape_sample_data_analysis, family = "binomial") 
landscape_model_3 <- glm(Window_859 ~ value + Cluster, data = landscape_sample_data_analysis, family = "binomial") 

summary(landscape_model) 
summary(landscape_model_2) # model 2 is best fit but NOT significant
summary(landscape_model_3)

r.squaredGLMM(landscape_model) 
r.squaredGLMM(landscape_model_2) # doesn't explain much variation
r.squaredGLMM(landscape_model_3)

car::Anova(landscape_model_2) %>%
  as.data.frame(row.names = NULL)

landscape_sample_data_analysis %>% 
  ggplot(aes(x = Window_859, y = value)) + 
  geom_boxplot()
```

**Window_859 - positively associated with p-a ratio - explaining 70% of variation**


```{r}
## Landscape percentage
landscape_sample_data_analysis <- landscape_sample_data %>% 
  filter(metric == "pland" & class_name == "Trees" & !is.na(n))

log_reg_output <- tibble("Estimate" = NA, "Std. Error" = NA, "z value" = NA, "Pr(>|z|)" = NA, "Window" = NA, "Variable" = NA)

for(i in levels(column_names$windows)){
  log_reg_output <- log_reg_output %>% 
    rbind(
      glm(get(i) ~ value * Cluster, data = landscape_sample_data_analysis, family = "binomial") %>%
        summary() %>%
        coef() %>%
        as.data.frame() %>%
        rownames_to_column(var = "Variable") %>%
        add_column(Window = paste0(i))
    )
  }

landscape_windows <- log_reg_output %>% 
  na.omit() %>%
  rename(P = "Pr(>|z|)") %>%
  arrange(P) %>%
  filter(Variable == "value")

# Subset to significant windows to reassess
landscape_model <- glm(Window_1196 ~ value * Cluster, data = landscape_sample_data_analysis, family = "binomial") 
landscape_model_2 <- glm(Window_1196 ~ value, data = landscape_sample_data_analysis, family = "binomial") 
landscape_model_3 <- glm(Window_1196 ~ value + Cluster, data = landscape_sample_data_analysis, family = "binomial") 

summary(landscape_model) 
summary(landscape_model_2) # model 1 is best fit but NOT significant
summary(landscape_model_3)

r.squaredGLMM(landscape_model) 
r.squaredGLMM(landscape_model_2) # doesn't explain much variation
r.squaredGLMM(landscape_model_3)

car::Anova(landscape_model_2) %>%
  as.data.frame(row.names = NULL)

landscape_sample_data_analysis %>% 
  ggplot(aes(x = Window_1196, y = anoph_density)) + 
  geom_boxplot()
```

**Close to significant for Window_1196 - but not quite - positive association with - only 20% of variation**



###########################################################################################################################################################################



# Presence/absence of introgressed window & anoph density 

## Mixed effects regression models - effect of anoph density on presence/absence of windows

```{r}
## Anoph density
log_reg_output <- tibble("Estimate" = NA, "Std. Error" = NA, "z value" = NA, "Pr(>|z|)" = NA, "Window" = NA, "Variable" = NA)

for(i in levels(column_names$windows)){
  log_reg_output <- log_reg_output %>% 
    rbind(
      glm(get(i) ~ anoph_density * Cluster, data = landscape_sample_data_analysis, family = "binomial") %>%
        summary() %>%
        coef() %>%
        as.data.frame() %>%
        rownames_to_column(var = "Variable") %>%
        add_column(Window = paste0(i))
    )
  }

anoph_windows <- log_reg_output %>% 
  na.omit() %>%
  rename(P = "Pr(>|z|)") %>%
  arrange(P) %>%
  filter(Variable == "anoph_density")

# Subset to significant windows to reassess
# 1504 - close but not really
landscape_model <- glm(Window_1504 ~ anoph_density * Cluster, data = landscape_sample_data_analysis, family = "binomial")
landscape_model_2 <- glm(Window_1504 ~ anoph_density, data = landscape_sample_data_analysis, family = "binomial") 
landscape_model_3 <- glm(Window_1504 ~ anoph_density + Cluster, data = landscape_sample_data_analysis, family = "binomial") 

summary(landscape_model) 
summary(landscape_model_2)
summary(landscape_model_3) # model 3 is best fit - not significant

r.squaredGLMM(landscape_model)
r.squaredGLMM(landscape_model_2)
r.squaredGLMM(landscape_model_3) # 90% of variation

car::Anova(landscape_model_3) %>%
  as.data.frame(row.names = NULL)

landscape_sample_data_analysis %>% 
  ggplot(aes(x = Window_1504, y = anoph_density)) + 
  geom_boxplot()

# 1236
landscape_model <- glm(Window_1236 ~ anoph_density * Cluster, data = landscape_sample_data_analysis, family = "binomial")
landscape_model_2 <- glm(Window_1236 ~ anoph_density, data = landscape_sample_data_analysis, family = "binomial") 
landscape_model_3 <- glm(Window_1236 ~ anoph_density + Cluster, data = landscape_sample_data_analysis, family = "binomial") 

summary(landscape_model) 
summary(landscape_model_2) # model 2 is best fit
summary(landscape_model_3)

r.squaredGLMM(landscape_model)
r.squaredGLMM(landscape_model_2) # but the variable only explains ~70% of the variance
r.squaredGLMM(landscape_model_3)

car::Anova(landscape_model_2) %>%
  as.data.frame(row.names = NULL)

landscape_sample_data_analysis %>% 
  ggplot(aes(x = Window_1236, y = anoph_density)) + 
  geom_boxplot()
```

**Window 1236 is negatively associated with anopheles density - only 4 samples**

# Map plots showing distribution of sampling and rasters

## Anopheles Layer 

```{r}
# Get malayis map data
malaysia <- ne_states(country = c("malaysia", "indonesia"), returnclass = "sf") %>% 
  mutate(name = ifelse(name != "Sabah" & name != "Sarawak", "Other", name))

# Plot Malaysia with raster layer and district points added
plot_1 <- ggplot() +
  geom_sf(data = malaysia, col = "transparent") +
  scale_x_continuous(limits = c(107, 120)) +
  scale_y_continuous(limits = c(-4.5, 7)) +
  geom_raster(data = anoph_raster_df, mapping = aes(x = x, y = y, fill = z)) +
  geom_sf(data = st_as_sf(polygon_data), col = "red", fill = "transparent") +
  scale_fill_viridis_c() +
  ylab("Latitude") +
  xlab("Longitude") +
  theme(legend.position = "none")

plot_2 <- ggplot() +
  geom_sf(data = malaysia, col = "transparent") +
  geom_raster(data = anoph_raster_df, mapping = aes(x = x, y = y, fill = z)) +
  geom_sf(data = st_as_sf(polygon_data), col = "red", fill = "transparent") +
  scale_x_continuous(limits = c(116, 117.5), breaks = c(116.5, 117)) +
  scale_y_continuous(limits = c(5.5, 7.5), breaks = c(6, 7)) +
  scale_fill_viridis_c() +
  ylab("Latitude") +
  xlab("Longitude") +
  labs(fill = "ALC Layer")

plot_raster <- grid.arrange(plot_1, plot_2, ncol = 2)
save_fig(plot_raster, "ALC_raster", width = 10, height = 6)

```

## Tree cover percentage layer & tree cover p:a ratio layer 
```{r,warning=F,message=F}
# Percentage
landscape_sample_data_analysis <- landscape_sample_data %>% 
  filter(metric == "pland" & class_name == "Trees" & !is.na(n))

polygon_data_landscape <- landscape_sample_data_analysis %>%
  dplyr::select(value, sampleid)  %>%
  rename(Percentage_TC = value) %>%
  left_join(sample_coords) %>%
  dplyr::select(x_gpsbackup, y_gpsbackup, sampleid, Percentage_TC) %>%
  pts2poly_centroids(0.0225, crs = 4326, agr = "constant") %>% # create polygons based on the 2.5km radius 
  vect() 

plot_1 <- ggplot() +
  geom_sf(data = malaysia, col = "transparent") +
  geom_sf(data = st_as_sf(polygon_data_landscape), col = "red", mapping = aes(fill = Percentage_TC)) +
  scale_x_continuous(limits = c(116, 117.5), breaks = c(116.5, 117)) +
  scale_y_continuous(limits = c(5.5, 7.5), breaks = c(6, 7)) +
  theme_dark() +
  scale_fill_viridis_c() +
  ylab("Latitude") +
  xlab("Longitude") 

# P:A ratio
landscape_sample_data_analysis <- landscape_sample_data %>% 
  filter(metric == "para_mn" & class_name == "Trees" & !is.na(n))

polygon_data_landscape <- landscape_sample_data_analysis %>%
  dplyr::select(value, sampleid)  %>%
  rename(PA_ratio_TC = value) %>%
  left_join(sample_coords) %>%
  dplyr::select(x_gpsbackup, y_gpsbackup, sampleid, PA_ratio_TC) %>%
  pts2poly_centroids(0.0225, crs = 4326, agr = "constant") %>% # create polygons based on the 2.5km radius 
  vect() 

plot_2 <- ggplot() +
  geom_sf(data = malaysia, col = "transparent") +
  geom_sf(data = st_as_sf(polygon_data_landscape), col = "red", mapping = aes(fill = PA_ratio_TC)) +
  scale_x_continuous(limits = c(116, 117.5), breaks = c(116.5, 117)) +
  scale_y_continuous(limits = c(5.5, 7.5), breaks = c(6, 7)) +
  theme_dark() +
  scale_fill_viridis_c() +
  ylab("Latitude") +
  xlab("Longitude") 

plot_raster <- grid.arrange(plot_1, plot_2, ncol = 2)
save_fig(plot_raster, "TC_raster", width = 10, height = 6)
```






```{r}
# Utility chunk (CSIRO app shp) — NOT a poster output. Disabled in migration:
#   - the Downloads example .shp was an absolute path to a throwaway str() check (dropped)
#   - st_write(append=TRUE) would mutate the imported raw malaysia.shp; eval=FALSE to keep raw read-only
# To regenerate malaysia.shp, re-enable and point st_write at data/processed/.
```

```{r, eval=FALSE}
# read_sf("~/Downloads/Example files/example_shp_poly/poly_10.shp") %>% str()  # dropped: exploratory, missing dep
ne_states(country = c("malaysia"), returnclass = "sf") %>%
  select(name) %>%
  filter(name == "Sabah") %>%
  st_cast("POLYGON") %>%
  rename(site_id = name) %>%
  rownames_to_column("name_1") %>%
  unite(site_id, 1:2) %>%
  st_write("data/processed/malaysia.shp", append=FALSE)
```



```{r, eval=FALSE}
# EXCLUDED in migration: this chunk read clinical/epi xlsx (PK_Sabah_Sample_naming_indexes
# + Epi_data_for_ZB_inventory) only to View() sample enrolment dates interactively.
# Output never assigned or used downstream, and the source files hold identifiable clinical
# data — intentionally not imported into the clean project. Re-enable only with a
# de-identified metadata source if enrolment dates are needed.
```


# Is there a relationship between genetic distance (IBD) and spatial distance

```{r}
library(geosphere)


dist_matrix <- distm(
  x = sample_coords %>% dplyr::select(x_gpsbackup, y_gpsbackup),
  fun = distHaversine  # You can use other distance measures (e.g., distVincentyEllipsoid)
)

rownames(dist_matrix) <- sample_coords$sampleid
colnames(dist_matrix) <- sample_coords$sampleid

dist_matrix <- dist_matrix %>%
  as.data.frame() %>% 
  rownames_to_column("sample1") %>%
  pivot_longer(2:ncol(.), names_to = "sample2", values_to = "distance") %>%
  mutate(pairwise = paste0(sample1, "-", sample2)) %>%
  dplyr::select(pairwise, distance) %>%
  left_join(
    read.table("data/raw/hmmIBD/Pk.hmm_fract_revisions.txt") %>%
      row_to_names(1) %>%
      mutate(sample1 = str_remove(sample1, "_DK.*"),
             sample2 = str_remove(sample2, "_DK.*")) %>%
      dplyr::select(sample1, sample2, fract_sites_IBD) %>%
      mutate(pairwise = paste0(sample1, "-", sample2)) %>%
      dplyr::select(pairwise, fract_sites_IBD) 
  ) %>% 
  na.omit() %>%
  mutate(fract_sites_IBD = as.numeric(fract_sites_IBD))

shapiro.test(dist_matrix$distance)
shapiro.test(dist_matrix$fract_sites_IBD)


cor.test(dist_matrix$distance, dist_matrix$fract_sites_IBD, method = "spearman")

glm(fract_sites_IBD ~ distance, data = dist_matrix) %>%
  summary()
```

